---
title: "align patterns"
author: "Nick Gauthier"
date: "7/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Aligning eofs from multiple sources

so first get some patterns
```{r}
n_modes <- 6

cera_patterns <- get_patterns(cera, k = n_modes, rotate = FALSE)
cesm_patterns <- get_patterns(filter(cesm, between(time, 1901, 2010)), k = n_modes, rotate = FALSE)
ccsm_patterns <- get_patterns(filter(ccsm, time >= 1901), k = n_modes, rotate = FALSE)
noaa_patterns <- get_patterns(filter(noaa, between(time, 1901, 2010)), k = n_modes, rotate = FALSE)
```

resample patterns to same resolution (probably the coarser of the two?)

cross correlate all the spatial patterns?

```{r}
plot(cera_patterns$eofs)
plot(ccsm_patterns$eofs)
```
```{r}
t1 <- as_tibble(cera_patterns$eofs) %>% 
  pivot_wider(names_from = PC, values_from = weight) %>% 
  select(-x,-y) %>% remove_missing()

t2 <- as_tibble(ccsm_patterns$eofs) %>% 
  pivot_wider(names_from = PC, values_from = weight) %>% 
  select(-x,-y) %>% remove_missing

cor(t1, t2, use = 'complete.obs')
```

```{r}
f1 <- as_tibble(st_crop(get_correlation(cera, cera_patterns),states_wus) ) %>% 
  pivot_wider(names_from = PC, values_from = SWE) %>% 
  select(-x,-y) %>% remove_missing()

f2 <- as_tibble(st_crop(get_correlation(ccsm, ccsm_patterns),states_wus) ) %>% 
  pivot_wider(names_from = PC, values_from = SWE) %>% 
  select(-x,-y) %>% remove_missing

library(corrplot)
corrplot(cor(f1, f2, use = 'complete.obs'))
```
```{r}
cor(f1, f2, use = 'complete.obs') %>% as_tibble(rownames = 'PC') %>% pivot_longer(-PC) %>%
  group_by(PC) %>%
  filter(abs(value) == max(abs(value)))

cor(f1, f2, use = 'complete.obs') %>% as_tibble(rownames = 'PC') %>% pivot_longer(-PC) %>%
  group_by(name) %>%
  filter(abs(value) == max(abs(value)))
```

```{r}
cos.sim=function(ma, mb){
  mat=tcrossprod(ma, mb)
  t1=sqrt(apply(ma, 1, crossprod))
  t2=sqrt(apply(mb, 1, crossprod))
  mat / outer(t1,t2)
}
cos.sim(as.matrix(f1), as.matrix(f2))
proxy::simil(f1, f2, method = 'cosine', by_rows = FALSE) %>% corrplot()
proxy::simil(t1, t2, method = 'cosine', by_rows = FALSE) %>% corrplot()
psych::factor.congruence(f1,f2) %>% corrplot()
psych::factor.congruence(t1,t2) %>% corrplot()
# the above are identical! so the burt/tucker index of factor congruence is the cosine similarity
```

```{r}
EFAtools::COMPARE(as.matrix(t1), as.matrix(t2), reorder = 'congruence', na.rm = TRUE)$diff_corres_cros
```
```{r}
psych::factor.congruence(t1,t2) %>%
  as_tibble(rownames = 'ref') %>%
  pivot_longer(-ref) %>%
  group_by(ref) %>%
  arrange(-abs(value), .by_group = TRUE) %>%
  filter(abs(value) >= 0.5)
```


so the problem with both correlation, pearson and spearman, and cosine similarities is that there is still mixing across the modes and the patterns aren't precisely matched!
rotate signs as needed


flag patterns with poor matching

visualize everything