---
title: "Downscaling snowpack variability in western North America"
subtitle: "Models and Observations"
author: "Nick Gauthier"
date: "Last knit on: `r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
link-citations: yes
---

Downscaling snow water equivalent 

# Introduction


Import the packages required for this analysis.

```{r setup}
knitr::opts_chunk$set(fig.width = 6, fig.asp = 0.618)

# analysis
library(remote) # empirical orthogonal teleconnections
library(telefit) # cca
library(tidyverse) # data manipulation and visualization
library(mgcv) # flexible nonlinear regression models
library(modelr)

# plotting
library(sf) # shapefiles and plotting
library(gganimate) # animated gifs
library(magick) # for multi-panel animations
library(scico) # color palettes
library(patchwork)


# this package
devtools::load_all()

load('../data.Rdata')
```
 
# Snowpack in the western US
 
First plot March SWE climatology from observations and reanalysis data. We immediately see the mismatch between the two because of the coarse resolution reanalysis. To drive home the point, here's the high-resolution observations, resampled to the reanalysis resolution. For a better comparison, each dataset was reduced to the common range of 1982--2010. The coarse data fail to capture the topographic heterogeneity of the domain, and thus the small-scale variations in elevation and temperature that influence snow accumulation and melt.

```{r, echo = FALSE, fig.cap="A) Mean March SWE from PRISM/SNOTEL observations, 1982-2017. B) Mean March SWE from CERA-20C Reanalysis, 1901-2010."}
a <- prism_dat %>%
  get_climatology() %>%
  ggplot() +
  geom_raster(aes(x, y, fill = swe_mean)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
   scale_fill_distiller(palette = 'Blues', direction = 1, name = 'SWE (mm)') +
  theme_void() +
  theme(legend.position = 'bottom')

b <- cera_dat %>%
  get_climatology() %>%
  ggplot() +
  geom_raster(aes(x, y, fill = swe_mean)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
   scale_fill_distiller(palette = 'Blues', direction = 1, name = 'SWE (mm)') +
  theme_void() +
  theme(legend.position = 'bottom')

a + b + plot_annotation(tag_levels = 'A')
```

When were the max and min SWE years?
```{r}
prism_dat %>%
  group_by(x,y) %>%
  filter(SWE == max(SWE)) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = year)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
   scale_fill_viridis_c() +
  theme_void() +
  theme(legend.position = 'bottom')

prism_dat %>%
  group_by(x,y) %>%
  filter(SWE == min(SWE)) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = year)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
     scale_fill_viridis_c() +
  theme_void() +
  theme(legend.position = 'bottom')
```

Where are the snowy pixels?
```{r}
prism_dat %>%
  group_by(x,y) %>%
  summarise(median_swe = median(SWE)) %>%
  ggplot() +
    geom_raster(aes(x, y, fill = median_swe > 3)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
   scale_fill_viridis_d() +
  theme_void() +
  theme(legend.position = 'bottom')
```


```{r}
prism_dat %>%
  ggplot(aes((SWE))) +
  geom_histogram()

prism_dat %>%
  group_by(x, y) %>%
  summarise(kurt =  e1071::kurtosis(SWE)) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = kurt > 3)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  theme(legend.position = 'bottom')

prism_dat %>%
  group_by(x, y) %>%
  summarise(skew =  e1071::skewness(SWE)) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = skew)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
scale_fill_viridis_c() +
  theme_void() +
  theme(legend.position = 'bottom')
  
prism_dat %>%
  group_by(x, y) %>%
  summarise(p = shapiro.test(log(SWE + 1))$p) %>%
  ggplot() +
    geom_raster(aes(x, y, fill = p < 0.05)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  scale_fill_viridis_d() 
```

```{r}
# prism_dat %>%
#   get_climatology() %>%
#   ggplot() +
#   geom_raster(aes(x, y, fill = swe_mean)) +
#   geom_sf(data = states_wus, fill = NA, color = 'black') +
#   scale_fill_scico(palette = 'davos', direction = -1, name = 'SWE (mm)') +
#   theme_void() +
#   theme(legend.position = 'bottom') +
#       ggtitle("Mean March Snow Water Equivalent", "1982-2017")
# ggsave('swe_mean.png', height = 7, width = 7)
a <- prism_dat %>%
  get_climatology() %>%
  ggplot() +
  geom_raster(aes(x, y, fill = swe_mean)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  scale_fill_scico(palette = 'davos', direction = -1, name = 'SWE (mm)') +
  theme_void() +
  theme(legend.position = 'bottom') +
      ggtitle("Mean March Snow Water Equivalent", "Observed")

b <- cera_dat %>%
  get_climatology() %>%
  ggplot() +
  geom_raster(aes(x, y, fill = swe_mean)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  scale_fill_scico(palette = 'davos', direction = -1, name = 'SWE (mm)') +
  theme_void() +
    ggtitle("", "Simulated") +
  theme(legend.position = 'bottom')

a + b + plot_annotation(tag_levels = 'A')
ggsave('swe_comparison.png', width = 7, height = .61*7)
```

Next we'll look at how SWE has varied over time and space.


# Modes of Snowpack Variability

In spite of the mismatch in mean values, perhaps the *variability* in SWE from year to year has something in common.

```{r animation, echo = FALSE, warning=FALSE,cache = TRUE}
# code courtesy of https://github.com/thomasp85/gganimate/wiki/Animation-Composition
c <- prism_dat %>%
      filter(between(year, 1990, 2010)) %>%
  group_by(x, y) %>%
  mutate(anomaly = (SWE - mean(SWE)) / sd(SWE))  %>% 
  ggplot() +
  geom_raster(aes(x, y, fill = anomaly)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  scale_fill_scico(palette = 'vik', direction = -1, limits = c(-4.5, 4.5), name = 'Standard\n deviations') +
    transition_states(year) +
    ggtitle("March {closest_state} SWE anomaly", "Observed") +
  theme_void() +
  theme(legend.position = 'bottom')

d <- cera_dat %>%
      filter(between(year, 1990, 2010)) %>%
    group_by(x, y) %>%
    mutate(anomaly = (SWE - mean(SWE)) / sd(SWE)) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = anomaly)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  scale_fill_scico(palette = 'vik', direction = -1, limits = c(-4.5, 4.5), name = 'Standard\n deviations') +
  transition_states(year) +
  ggtitle("", "Simulated") +
  theme_void() +
  theme(legend.position = 'bottom')

c_gif <- gganimate::animate(c)
d_gif <- gganimate::animate(d)

c_mgif <- image_read(c_gif)
d_mgif <- image_read(d_gif)

new_gif <- image_append(c(c_mgif[1], d_mgif[1]))
for(i in 2:100){
  combined <- image_append(c(c_mgif[i], d_mgif[i]))
  new_gif <- c(new_gif, combined)
}

new_gif

image_write(new_gif, 'anomalies.mp4')
```

## PCA

We can decompose all this year-to-year variability into the dominant spatial modes -- robust spatial anomaly patterns that tend to co-occur.

Run a principal components analysis. Get the eigenvalues, and run some truncation tests.

```{r, fig.width = 8, fig.asp = .4, echo = FALSE}
e <- prism_dat %>% 
  get_pcs() %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 10) +
  scale_color_brewer(palette = 'Spectral')

f <- cera_dat %>% 
  get_pcs() %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 10) +
  scale_color_brewer(palette = 'Spectral')

e + f + plot_annotation(tag_levels = 'A')
ggsave('eigenvalues.png', width = 8, height = 3.2)
```

```{r}
dof_p <- prism_dat %>% 
  get_pcs() %>%
  get_eigenvalues() %>%
  pull(eigenvalues)

sum(dof_p)^2 / sum(dof_p^2) # from bretheron et al 1999

dof_c <- cera_dat %>% 
  get_pcs() %>%
  get_eigenvalues() %>%
  pull(eigenvalues)

sum(dof_c)^2 / sum(dof_c^2) # from bretheron et al 1999
```

```{r}
# prism for the whole period
 prism_dat %>% 
  get_pcs() %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 10) +
  scale_color_brewer(palette = 'Spectral')

# prism for before 2010
 prism_dat %>% 
   filter(year <=2010)%>%
  get_pcs() %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 10) +
  scale_color_brewer(palette = 'Spectral')

 #cera after 1983
cera_dat %>% 
  filter(year>=1983) %>%
  get_pcs() %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 10) +
  scale_color_brewer(palette = 'Spectral')

#cera whole
cera_dat %>% 
  get_pcs() %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 10) +
  scale_color_brewer(palette = 'Spectral')
```

Looks like there are ~4 spatial degrees of freedom in the data

## EOFs

So let's stick with the 4 and 4 solution, and investigate those patterns more. 
Lets start by truncating at 4. Extract the pc amplitude time series and EOF spatial patterns for the leading 4 modes.

```{r}
n_modes <- 5
```

```{r}
prism_patterns <- get_patterns(prism_dat, k = n_modes)
cera_patterns <- get_patterns(cera_dat, k = n_modes)
```


```{r, echo = FALSE}
plot_amps(prism_patterns) + geom_hline(yintercept = 0, linetype = 2) 
plot_amps(cera_patterns) + geom_hline(yintercept = 0, linetype = 2) 
```

Plot the eofs. 
```{r echo = FALSE, fig.asp = 1}
plot_eofs(prism_patterns, 'vik', scaled = FALSE) 
plot_eofs(cera_patterns, 'vik', scaled = FALSE)
plot_eofs(prism_patterns, 'vik', scaled = TRUE) 
plot_eofs(cera_patterns, 'vik', scaled = TRUE)
```

```{r, fig.width = 8, fig.asp = .7}
prism_patterns$eofs_corr <- prism_patterns$eofs_corr %>% mutate(correlation = if_else(PC == 4, correlation, correlation * -1))

a<-plot_eofs(prism_patterns, 'vik', scaled = TRUE) + facet_wrap(~paste0('EOF', PC), nrow = 1) + ggtitle('Leading March SWE spatial modes', 'Observed') + theme(legend.position = 'bottom')

b<-plot_eofs(cera_patterns, 'vik', scaled = TRUE) + facet_wrap(~paste0('EOF', PC), nrow = 1) + ggtitle('', 'Simulated') + theme(legend.position = 'bottom')

c <- a / b & theme(legend.position = "bottom")
c + plot_layout(guides = "collect", heights = 1,1)
ggsave('eofs.png')
```

```{r}
# For each PC amplitude time series, plot out the wettest and driest years from the observations.
# From this we can tell that pc1 and 2 need to reverse signs to be physically consistent
extreme_years <- prism_patterns$amplitudes %>%
  group_by(PC) %>%
  filter(amplitude == max(amplitude) | amplitude == min(amplitude)) %>%
  pull(year)

(prism - mean(prism)) %>%
  .[[extreme_years - 1981]] %>%
  setNames(extreme_years) %>%
  as.data.frame(xy = TRUE, na.rm = TRUE) %>%
  gather(year, swe, 3:10) %>%
  mutate(year = factor(year, levels = paste0('X', extreme_years))) %>%
  ggplot(aes(x, y, fill = swe)) +
  geom_raster() +
  facet_wrap(~year) +
  coord_quickmap() +
  scale_fill_distiller(palette = 'BrBG', direction = 1, limits = c(-1500, 1500)) +
  theme_void()
```

## Internal variability

```{r, fig.asp = .9}
prism_recon <- reconstruct_field(prism_patterns)
```

look at the mean and standard deviation of the unexplained variance. We're only focusing on "snowy pixels" where the median is > 3mm
```{r}
resids <- prism_recon %>%
   left_join(prism_dat, by = c('x', 'y', 'year'), suffix = c('_recon', '_obs')) %>%
    mutate(error = SWE_recon - SWE_obs) %>%
  group_by(x, y) %>%
  filter(median(SWE_recon) > 3 & median(SWE_obs) > 3) %>%
  ungroup()

resids %>%
  group_by(x, y) %>%
  summarise(avg = mean(error),
            stdev = sd(error)) %>%
    ggplot() +
  geom_raster(aes(x, y, fill = avg)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  scale_fill_viridis_c() +
  ggtitle('Error from dimensionality reduction', 'PRISM')
```

```{r}
prism_recon %>%
   get_errors %>%
  group_by(x, y) %>%
   # filter(median(SWE_recon) > 3 & median(SWE_obs) > 3) %>%
  summarise(avg = mean(accuracy_ratio)) %>%
    ggplot() +
  geom_raster(aes(x, y, fill = avg)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  scale_fill_viridis_c() +
  ggtitle('Error from dimensionality reduction', 'PRISM')
```

Differences in snowy pixels. in general, by truncating at 4, we slightly overestimate the amount of snowy pixels.
```{r}
prism_recon %>%
   left_join(prism_dat, by = c('x', 'y', 'year'), suffix = c('_recon', '_obs'))%>%
  group_by(x, y) %>%
  summarise(rec_med = median(SWE_recon),
            obs_med = median(SWE_obs)) %>%
  mutate(obs_snowy = obs_med > 3,
         rec_snowy = rec_med > 3,
         agreed = case_when(obs_snowy == rec_snowy & obs_snowy == TRUE ~'True positive',
                            obs_snowy == rec_snowy & obs_snowy == FALSE ~'True negative',
                            obs_snowy != rec_snowy & obs_snowy == TRUE ~'False negative',
                            obs_snowy != rec_snowy & obs_snowy == FALSE ~ 'False positive')) %>%
      ggplot() +
  geom_raster(aes(x, y, fill = agreed)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  scale_fill_viridis_d()
```


Look here, the observations have some extremely (suspect?) high values -- noise that the pca removes? or real?
```{r}
resids %>%
  ggplot(aes(SWE_obs, SWE_recon)) +
  geom_point(alpha = .1) +
  geom_smooth() +
  theme_bw()

resids %>%
  ggplot(aes(error)) +
  geom_histogram() +
  theme_bw()

resids %>% group_by(x, y) %>%
  filter(median(SWE_recon) > 3 & median(SWE_obs) > 3) %>%
  ggplot(aes(error)) +
  geom_histogram() +
  theme_bw()
```

```{r}
resids %>%
  arrange(error)
```


Is it normal?


```{r}
resids %>%
  group_by(x, y) %>%
  summarise(p = shapiro.test(error)$p) %>%
  ggplot() +
    geom_raster(aes(x, y, fill = p < 0.05)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  scale_fill_viridis_d() +
  ggtitle('Error from dimensionality reduction', 'PRISM')
```


```{r}
score_map <- prism_recon %>%
  get_errors() %>%
  group_by(x, y) %>%
  get_scores()

score_ts <- prism_recon %>%
  get_errors() %>%
    group_by(x, y) %>%
    filter(median(SWE_recon) > 3 & median(SWE_obs) > 3) %>%
  ungroup() %>%
  group_by(year) %>%
  get_scores()
```


```{r, fig.width = 10}
score_ts %>%
  gather(metric, value, me:rmsle) %>%
  ggplot(aes(year, value)) +
  geom_line() +
  facet_wrap(~metric, scales = 'free_y') +
  theme_bw()
```


```{r}
  plot_fun <- function(x, name) {
      ggplot(x) +
  geom_raster(aes(x, y, fill = value)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  scale_fill_viridis_c(name = name)
  }

tmp <- score_map %>%
gather(metric, value, me:rmsle) %>%
group_by(metric) %>%
  nest %>%
  mutate(plots = map2(data, metric, plot_fun))
tmp$plots
```

```{r}
score_map %>%
      ggplot() +
  geom_raster(aes(x, y, fill = srmse < .7)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  scale_fill_viridis_d()
```


## Trends

Are there any trends?

```{r}
trends_mod <- prism_patterns$amplitudes %>%
  mutate(PC = paste0('PC', PC)) %>% 
  group_by(PC) %>%
  nest() %>%
  mutate(mod = map(data, ~gam(amplitude ~ 
                                s(year, k = 4), 
                              data = ., 
                              method = 'REML', 
                              select = TRUE))) %>%
  mutate(r2 = map_dbl(mod, ~summary(.)$r.sq)) %>%
  mutate(data = map2(data, mod, add_predictions),
         data = map2(data, mod, add_residuals)) 

walk(trends_mod$mod, plot)
```

```{r}

test <- prism_dat %>%
  group_by(x, y) %>%
  nest() %>%
  mutate(mod = map(data, ~lm(SWE ~ year, data = .) %>% broom::tidy())) %>%
  select(-data) %>%
  unnest %>%
  filter(term == 'year')


test %>%
   ggplot() +
    geom_raster(aes(x, y, fill = estimate)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  scale_fill_distiller(palette = 'RdBu')


test %>%
   ggplot() +
    geom_raster(aes(x, y, fill = p.value < .05)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() 
```


# Coupled Pattern Analysis

First look at just the error from the dimensionality reduction  with the PCs. This is basically the error in the pc targets, so is the minimum error we'd get even from the best downscaling

## GLM

```{r}
library(glmnet)
library(MuMIn)

gam_formula <-  prep_data(cera_patterns, prism_patterns)  %>%
    # this should be more flexible below, checking for innapropriate predictors
    dplyr::select(-PC, -amplitude, -year) %>%
    names() %>%
    map( ~ paste0("s(", ., ", bs = 'cr', k = 4)")) %>%
    paste(collapse = ' + ') %>%
    paste('amplitude ~ ', .) %>%
    as.formula
glm_formula <- prep_data(cera_patterns, prism_patterns)  %>%
    # this should be more flexible below, checking for innapropriate predictors
    dplyr::select(-PC, -amplitude, -year) %>%
    names() %>%
    paste(collapse = ' + ') %>%
    paste('amplitude ~ ', .) %>%
    as.formula

test <- prep_data(cera_patterns, prism_patterns) %>%
   group_by(PC) %>%
    nest()# %>%
testdat <- test[[1,2]][[1]]

glm(glm_formula, data = testdat)

t1 <- lm(glm_formula, data = testdat, na.action = "na.fail")

test_gam <- gam(gam_formula, method = 'REML',select = TRUE, data = testdat, na.action = "na.fail")
test_gam2 <- gam(gam_formula, data = testdat, na.action = "na.fail")

t2 <- dredge(t1)
summary(get.models(t2, 1)[[1]])
t2

t3 <- dredge(test_gam2)
t3
summary(get.models(t3, 1)[[1]])

plot(test_gam)

t4 <- glmnet(testdat %>%
  select(PC1:PC9) %>% 
  as.matrix(), pull(testdat, amplitude), standardize = FALSE)
plot(t4)
coef(t4, s = .1);coef(t4, s = .05)

# this is unstable
t5 <- cv.glmnet(testdat %>%
  select(PC1:PC9) %>% 
  as.matrix(), pull(testdat, amplitude), standardize = FALSE, nfolds = 5)
plot(t5)
coef(t5, s = "lambda.1se")

# this isnt!
t6 <- cv.glmnet(testdat %>%
  select(PC1:PC9) %>% 
  as.matrix(), pull(testdat, amplitude), standardize = FALSE, nfolds = 5, foldid = rep(1:5, each = 6, length.out =29))
plot(t6)
coef(t6, s = "lambda.1se")
coef(t6, s = "lambda.min")

t7<- prep_data(cera_patterns, prism_patterns) %>%
   group_by(PC) %>%
    nest() %>%
mutate(mod1 = map(data, ~ gam(gam_formula, data = ., method = 'REML',select = TRUE)),
       mod2 = map(data, ~ eval(getCall(dredge(gam(gam_formula, data = ., na.action = "na.fail")), 1))),
       mod3 = map(data, ~ eval(getCall(dredge(lm(glm_formula, data = ., na.action = "na.fail")), 1)))) %>%
    ungroup() %>%
    mutate(r21 = map_dbl(mod1, ~ sqrt(mean(.$residuals^2))),
           r22 = map_dbl(mod2, ~ sqrt(mean(.$residuals^2))),
           r23 = map_dbl(mod3, ~ sqrt(mean(.$residuals^2)))) 

sqrt(mean(res$residuals^2))

t7$mod1[[1]] %>% plot
t7$mod2[[1]] %>% plot
```


## GAM

Let's look under the hood of what the GAM models above were doing.
```{r}
gam_mod <- couple_patterns(cera_patterns, prism_patterns)
```

```{r}
walk(gam_mod$model$mod, plot)
```

How do the models perform?
```{r}
gam_mod$model %>%
  select(PC, r2) %>%
  arrange(-r2)
```

Are the residuals autocorrelated? Some are
```{r}
gam_mod$data %>%
  group_by(PC) %>%
  summarise(autocor = cor(resid, lag(resid), use = 'pairwise.complete.obs')) %>%
  arrange(-abs(autocor))
```

```{r}
ggplot(gam_mod$data, aes(resid)) +
  #geom_histogram() + 
  geom_density(fill = NA) +
  geom_vline(xintercept = 0, color = 'red', linetype = 2) +
  facet_wrap(~PC) +
  theme_bw()
```

```{r}
gam_recon <- predict_patterns(gam_mod, cera_patterns) %>%
  reconstruct_field(prism_patterns, .)

gam_score_map <- gam_recon %>%
  get_errors() %>%
  group_by(x, y) %>%
  get_scores()
```


```{r}
gam_score_map %>%
 # group_by(x, y) %>%
#  summarise(corr = cor(SWE_recon, SWE_obs)) %>%
   ggplot() +
      geom_raster(aes(x, y, fill = corr)) +
      geom_sf(data = states_wus, fill = NA, color = 'black') +
      scale_fill_viridis_c() +
      theme_void()
```

```{r}
ggplot(gam_score_map ) +
      geom_raster(aes(x, y, fill = srmse < .7)) +
      geom_sf(data = states_wus, fill = NA, color = 'black') +
      scale_fill_viridis_d() +
      theme_void()
```
```{r}
ggplot(gam_score_map ) +
      geom_raster(aes(x, y, fill = srmse)) +
      geom_sf(data = states_wus, fill = NA, color = 'black') +
      scale_fill_viridis_c() +
      theme_void()
```


```{r}
e <- prism_dat %>%
      filter(between(year, 1990, 2010)) %>%
  group_by(x, y) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = SWE)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  scale_fill_scico(palette = 'davos', direction = -1, name = 'SWE (mm)') +
    transition_states(year) +
    ggtitle("March {closest_state} SWE", "Observed") +
  theme_void() +
  theme(legend.position = 'bottom')

f <- gam_recon %>%
      filter(between(year, 1990, 2010)) %>%
  group_by(x, y) %>%
  mutate(anomaly = (SWE - mean(SWE)) / sd(SWE))  %>% 
  ggplot() +
  geom_raster(aes(x, y, fill = SWE)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  scale_fill_scico(palette = 'davos', direction = -1, name = 'SWE (mm)') +
    transition_states(year) +
    ggtitle("", "Reconstructed") +
  theme_void() +
  theme(legend.position = 'bottom')


e_gif <- gganimate::animate(e)
f_gif <- gganimate::animate(f)

e_mgif <- image_read(e_gif)
f_mgif <- image_read(f_gif)

new_gif <- image_append(c(e_mgif[1], f_mgif[1]))
for(i in 2:100){
  combined <- image_append(c(e_mgif[i], f_mgif[i]))
  new_gif <- c(new_gif, combined)
}

new_gif

image_write(new_gif, 'reconstruction.gif')
```


```{r}
knitr::knit_exit()
```

## CCA

```{r}
cera_can <- cera_patterns$eofs %>%
  select(-value) %>%  # use normalized or unnormalized?
  spread(EOF, weight) %>%
  select(-x, -y) %>%
  as.matrix() %>%
  `%*%`(cca$ycoef) %>%
    `colnames<-`(paste0('PC', 1:n_modes)) %>%
  as_tibble()
  
prism_can <- prism_patterns$eofs %>%
  select(-value) %>%
  spread(EOF, weight) %>%
  select(-x, -y) %>%
  as.matrix() %>%
  `%*%`(cca$xcoef) %>%
  `colnames<-`(paste0('PC', 1:n_modes)) %>%
  as_tibble()
```

```{r}
cera_dat %>%
  spread(year, SWE) %>%
  mutate(column = 1:n()) %>%
  select(x, y, column) %>%
  bind_cols(cera_can) %>%
  gather(pattern, value, starts_with('PC')) %>% 
  ggplot() +
  geom_raster(aes(x, y, fill = value)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~pattern) +
      scale_fill_scico(palette = 'broc', direction = -1, limits = c(-1, 1) * 8) +
  theme_void()+
  coord_sf() +
  ggtitle('Reanalysis CCA')

prism_dat %>%
  spread(year, SWE) %>%
  mutate(column = 1:n()) %>%
  select(x, y, column) %>%
  bind_cols(prism_can) %>%
  gather(pattern, value, starts_with('PC')) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = value)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~pattern) +
      scale_fill_scico(palette = 'broc', direction = -1, limits = c(-1, 1) * 70) +
  theme_void()+
  coord_sf() +
  ggtitle('Observed CCA')
```

```{r}
cera_can <- cera_patterns$eofs %>%
  select(-value) %>%  # use normalized or unnormalized?
  spread(EOF, weight) %>%
  select(-x, -y) %>%
  as.matrix() %>%
  `%*%`(cca$ycoef) %>%
    `colnames<-`(paste0('PC', 1:n_modes)) %>%
  as_tibble()

cera_patterns$amplitudes  

prism_can <- prism_patterns$eofs %>%
  select(-value) %>%
  spread(EOF, weight) %>%
  select(-x, -y) %>%
  as.matrix() %>%
  `%*%`(cca$xcoef) %>%
  `colnames<-`(paste0('PC', 1:n_modes)) %>%
  as_tibble()
```


```{r}
# make belwo into a function
t1 <- prism_dat %>%
    group_by(x,y) %>%
  mutate(test = all(SWE == 0)) %>%
  filter(test == FALSE) %>%
  select(-test) %>%
  ungroup() %>%
  spread(year, SWE) %>%
  select(-x, -y) %>%
  as.matrix()

t2 <- cera_dat %>%
  filter(year >= 1982) %>% 
  spread(year, SWE) %>%
  select(-x, -y) %>%
  as.matrix()

tele.preds <- cca.predict(t2, t1, t2, k.x = 5, k.y = 5)
1 - var(as.numeric(tele.preds-t1)) / var(as.numeric(t1))

```

it looks like for some reason the cca.predict function isn't rescaling correctly? or its just meant to work with standardized anomalies?

```{r}
prism_scaler <- prism_dat %>%
  group_by(x,y) %>%
  summarise(avg = mean(SWE), sdev = sd(SWE, na.rm = TRUE))
```

```{r}
prism_dat %>%
  filter(year <= 2010) %>%
    group_by(x,y) %>%
  mutate(test = all(SWE == 0)) %>%
  filter(test == FALSE) %>%
  select(-test) %>%
    ungroup() %>%
  spread(year, SWE) %>%
  select(x, y) %>%
  cbind(tele.preds) %>%
  gather(year, SWE, `1982`:`2010`) %>%
  left_join(prism_scaler) %>%
  mutate(SWE = SWE * sdev + avg) %>%
  filter(year <=1987) %>%
  ggplot(aes(x,y)) +
  geom_raster(aes(fill = SWE)) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_quickmap()
```

```{r}
prism_dat %>%
  filter(year <=1987) %>%
  ggplot(aes(x,y)) +
  geom_raster(aes(fill = SWE)) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_quickmap()
```


```{r}
anomaly_dat <- prism_dat %>%
  filter( year <= 2010) %>% # important to make sure everything's calculated on the same time scale, may be best to do this further up?
  group_by(x, y) %>%
  mutate(swe_mean = mean(SWE),
         swe_sd = sd(SWE, na.rm = TRUE),
         anomaly = SWE - swe_mean) %>%
  ungroup()

tele_dat <- tele.preds %>%
  as_tibble() %>%
  bind_cols(t3, .) %>%
  gather(year, tele_pred, `1982`:`2010`) %>%
  mutate(year = as.numeric(year)) %>%
  left_join(anomaly_dat)  %>%
  mutate(swe_pred = tele_pred * swe_sd + swe_mean,
         res = swe_pred - SWE) 
```

```{r, echo = FALSE}
tele_dat %>%
  filter(year < 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = tele_pred)) +
  scale_fill_viridis_c() +
  facet_wrap(~year) +
  coord_quickmap()

tele_dat %>%
  filter(year < 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = anomaly)) +
  scale_fill_distiller(palette = 'RdBu') +
  facet_wrap(~year) +
  coord_quickmap()
```

```{r, echo = FALSE}
tele_dat %>%
  filter(year == 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = swe_mean)) +
  scale_fill_viridis_c() +
  facet_wrap(~year) +
  coord_quickmap()

tele_dat %>%
  filter(year == 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = swe_sd)) +
  scale_fill_viridis_c() +
  facet_wrap(~year) +
  coord_quickmap()
```

```{r}
plot(mean(prism));plot(calc(prism, sd))
```

```{r}
tele_dat %>%
  mutate(test = tele_pred * swe_sd + swe_mean) %>%
    filter(year == 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = test)) +
  scale_fill_viridis_c() +
  facet_wrap(~year) +
  coord_quickmap()

tele_dat %>%
  mutate(test = tele_pred * swe_sd + swe_mean) %>%
    filter(year == 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = SWE)) +
  scale_fill_viridis_c() +
  facet_wrap(~year) +
  coord_quickmap()
```

```{r}
tele_dat %>%
  mutate(swe_pred = tele_pred * swe_sd + swe_mean,
         error = swe_pred - SWE) %>%
      filter(year < 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = error)) +
  scale_fill_distiller(palette = 'RdBu') +
  facet_wrap(~year) +
  coord_quickmap()
```

## EOT

```{r}
cera_rast <- cera_dat %>%
  filter(year >= 1982) %>%
  spread(year, SWE) %>%
  rasterFromXYZ() 

prism_rast <- prism_dat %>%
  filter(year <= 2010) %>%
  spread(year, SWE) %>%
  rasterFromXYZ()

eot1 <- cera_rast %>%
  anomalize() #%>%
 # denoise(expl.var = 0.9, weighted = TRUE)

eot2 <- prism_rast %>%
  anomalize()# %>% 
  #denoise(k = 5, weighted = TRUE)
```

```{r}
telecon <- eot(eot1, eot2, n = 4, standardised = FALSE)
```


```{r}
walk(1:4, ~plot(telecon, y = ., show.bp = TRUE))
```

```{r}
predict(telecon, eot1, n = 4) %>% .[[1]] %>% `+`(mean(prism_rast))%>% plot
```

cv
```{r}
cera_rast <- cera_dat %>%
  filter(year >= 1982) %>%
  spread(year, SWE) %>%
  rasterFromXYZ() 

prism_rast <- prism_dat %>%
  filter(year <= 2010) %>%
  spread(year, SWE) %>%
  rasterFromXYZ()

eot1 <- cera_rast %>%
  anomalize()# %>%
  #denoise(k = 5, weighted = TRUE)

eot2 <- prism_rast %>%
  anomalize() #%>% 
  #denoise(k = 5, weighted = TRUE)

telecon <- eot(eot1, eot2, n = 4, standardised = FALSE, verbose = FALSE)

predict(telecon, eot1, n = 4) + mean(prism_rast)

fit_kfold_eot <- function(k_preds, k_obs, preds, obs){
  # this should be refactored ...
  # find the years of overlap between the response and predictor fields
  years <- intersect(preds$year, obs$year)
  preds <- filter(preds, year %in% years)
  obs <- filter(obs, year %in% years)

  # divide years into 5 contiguous folds
  n <- length(years)
  r <- n %% 5
  fold_times <- rep(n %/% 5, 5)
  fold_times[1:r] <- fold_times[1:r] + 1
  folds <- tibble(year = years, fold = rep(1:5, times = fold_times))

  # preprocess the training data for each fold
  obs_train_rast <- purrr::map(1:5, ~ filter(folds, fold == .) %>%
                     anti_join(obs, ., by = 'year') %>% # training years
                       spread(year, SWE) %>%
                      rasterFromXYZ())
  
  obs_train <- purrr::map(obs_train_rast, ~ anomalize(.))# %>% denoise(expl.var = 0.9, weighted = TRUE, verbose = FALSE))

  pred_train <- purrr::map(1:5, ~ filter(folds, fold == .) %>%
                      anti_join(preds, ., by = 'year') %>% # training years
                        spread(year, SWE) %>%
                      rasterFromXYZ() %>%
                        anomalize) #%>%
                        #denoise(expl.var = 0.9, weighted = TRUE, verbose = FALSE))

  # preprocess test data for each fold
  test <- purrr::map(1:5, ~ filter(folds, fold == .) %>%
                semi_join(preds, ., by = 'year') %>%
                  spread(year, SWE) %>%
                      rasterFromXYZ()) # test years

  # fit model to training data and use to predict new fields
  list(eot = map2(pred_train, obs_train, ~eot(.x, .y, n = 6, standardised = FALSE, verbose = TRUE)),
       test = test,
       mean = purrr::map(obs_train_rast, mean))
}

cv_eot <- expand_grid(k_predictor = 1:9, k_response = 1:9) %>%
  mutate(cv = map2(k_predictor, k_response, ~fit_kfold_eot(.x, .y, preds = cera_dat, obs = prism_dat)))



cv_eot <- fit_kfold_eot(1,1, preds = cera_dat, obs = prism_dat)

map(cv_eot, predict)

map(cv_eot, ~nXplain(.,.73))
purrr::map(cv_eot$eot, ~plot(., y = 3, show.bp = TRUE))

cv_eot$test %>% brick %>% plot

cv_eot$mean %>% brick() %>% plot

for(i in 1:6) {
  eot_pred <- pmap(list(cv_eot$eot, cv_eot$test, cv_eot$mean), ~  predict(..1, ..2, n = i) + ..3) %>%
  brick

cellStats(sqrt(mean((eot_pred - prism_rast)^2)), mean) %>% print
}


test_cv <- fit_kfold(5, 5, cera_dat, prism_dat)

test_cv %>% get_errors() %>% pull(error) %>% .^2 %>% mean %>% sqrt

(predict(cv_eot$eot[[1]], cv_eot$test[[1]], n = 1) + cv_eot$mean[[1]]) %>% plot

(predict(cv_eot$eot[[1]], cv_eot$test[[1]], n = 3) + cv_eot$mean[[1]]) %>% plot
(predict(cv_eot$eot[[1]], cv_eot$test[[1]], n = 6) + cv_eot$mean[[1]]) %>% plot
plot(prism_rast[[1:6]])
test_cv %>%
  filter(year <= 1986) %>%
  select(-anomaly) %>%
  spread(year, SWE) %>%
  rasterFromXYZ() %>% plot

```

## Delta

```{r}
delta <- function(pred, obs, newdata) {
  
  pred_clim <- get_climatology(pred)
  obs_clim <- get_climatology(obs)
  
  newdata %>%
    left_join(pred_clim, by = c('x', 'y')) %>%
  mutate(anomaly = SWE / swe_mean) %>%
    select(x,y, year, anomaly) %>%
    group_by(year) %>%
    nest() %>%
    mutate(mod = map(data, ~ gam(anomaly ~ s(x, y, k = 300, bs = 'gp'), data = .))) %>%
    select(-data) %>%
    mutate(data = rep(list(obs_clim), n())) %>%
    mutate(anomaly = map2(mod, data, predict, type = 'response')) %>%
    select(-mod) %>%
    unnest(c(data, anomaly)) %>% 
    mutate(SWE = swe_mean * if_else(anomaly >=0, anomaly, 0)) %>%
    select(year, x, y, SWE)
}

prep_kfold_delta <- function(preds, obs, type = NULL){
    # find the years of overlap between the response and predictor fields
  years <- intersect(preds$year, obs$year)
  preds <- filter(preds, year %in% years)
  obs <- filter(obs, year %in% years)

  # divide years into 5 contiguous folds
  n <- length(years)
  r <- n %% 5
  fold_times <- rep(n %/% 5, 5)
  fold_times[1:r] <- fold_times[1:r] + 1

  folds <- tibble(year = years, fold = rep(1:5, times = fold_times))

  train_obs <- map(1:5, ~ filter(folds, fold == .) %>%
                     anti_join(obs, ., by = 'year'))

  train_preds <- map(1:5, ~ filter(folds, fold == .) %>%
                      anti_join(preds, ., by = 'year'))

  # preprocess test data for each fold
  test <- map(1:5, ~ filter(folds, fold == .) %>%
                semi_join(preds, ., by = 'year')) # test years

  tibble(train_obs, train_preds, test)
}

t2 <- prep_kfold_delta(cera_dat, prism_dat)
 
t3 <- pmap_dfr(list(t2$train_preds, t2$train_obs, t2$test), delta)

t4 <- t3 %>% get_errors() %>%
    left_join(areas, by = c("x", "y")) %>%
    group_by(year) %>%
    summarise(SWE_obs = sum(SWE_obs * area),
              SWE_recon = sum(SWE_recon * area)) %>%
    mutate(error = SWE_recon - SWE_obs)
```

```{r}

t1 <- delta(filter(cera_dat, year >= 1982), filter(prism_dat, year <= 2010), filter(cera_dat, year == 1991))


 ggplot(t1) +
  geom_raster(aes(x, y, fill = SWE)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  scale_fill_viridis_c(limits = c(0, 3500)) +
  theme_void() +
  theme(legend.position = 'bottom')
 
  ggplot(prism_dat %>% filter(year == 1991)) +
  geom_raster(aes(x, y, fill = SWE)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  scale_fill_viridis_c(limits = c(0, 3500)) +
  theme_void() +
  theme(legend.position = 'bottom')
  
  
  predict_cca(cera_dat %>% filter(year >= 1982)%>% get_patterns(k = 4), prism_dat %>% filter(year <= 2010) %>% get_patterns(k = 4), cera_dat %>% filter(year == 1991)) %>% 
     ggplot() +
  geom_raster(aes(x, y, fill = SWE)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  scale_fill_viridis_c(limits = c(0, 3500)) +
  theme_void() +
  theme(legend.position = 'bottom')


```
so what above must change by folds? just the means for each training period? -- but wait, the newdata here is the observations from the test period

# Validation


### Cross-validation

Set up a 5 fold cross validation experiment for selecting the truncation level, $k$, for each set of PCs.


```{r}
cv_dat2 <- expand_grid(k_predictor = 1:9, k_response = 1:9) %>%
  mutate(kfolds = map2(k_predictor, k_response, ~prep_kfold(.x, .y, preds = cera_dat, obs = prism_dat)),
         gam = map(kfolds, fit_cv, fun = predict_gam),
         cca = map(kfolds, fit_cv, fun = predict_cca)) %>%
  select(-kfolds) %>%
  gather(type, metric, gam:cca) 

saveRDS(cv_dat, 'cv_dat')
```

```{r}
cv_dat2%>%
  gather(type, metric, gam:cca) %>%
  unnest(metric) %>%
  ggplot(aes(year, group = interaction(k_predictor, k_response, type)))+
    geom_line(aes(y = SWE_obs), color = 'red') +
  geom_line(aes(y = SWE_recon, color = type), alpha = .1) 
```

```{r}
cv_dat2 %>%
  gather(type, metric, gam:cca)%>%
  mutate(rmse = map_dbl(metric, ~sqrt(mean((.$SWE_recon - .$SWE_obs)^2)))) %>%
  select(-metric) %>%
  arrange(rmse)  %>%
  filter(rmse <= (min(rmse) + sd(rmse))) %>%
      mutate(test = k_predictor + k_response) %>%
  arrange(test)
```

```{r}
t4 %>%
  summarise(rmse = sqrt(mean((SWE_recon - SWE_obs)^2)))
```


```{r}
cv_dat2 %>%
  gather(type, metric, gam:cca)%>%
  mutate(rmse = map_dbl(metric, ~sqrt(mean((.$SWE_recon - .$SWE_obs)^2)))) %>%
  select(-metric) %>%
      ggplot(aes(k_predictor, k_response, fill = rmse)) +
  geom_tile() +
  scale_fill_viridis_c(direction = -1) +
  facet_wrap(~type) +
  scale_x_continuous(breaks = 1:9, minor_breaks = NULL) +
  scale_y_continuous(breaks = 1:9, minor_breaks = NULL) +
  coord_equal() +
  theme_minimal()
```



```{r}
loadRDS('cv_dat')
cv_dat %>%
  gather(type, metric, gam:cca) %>%
  mutate(rmse = map_dbl(metric, ~mean(.$rmse))) %>%
  select(-metric) %>%
  arrange(rmse) %>%
  filter(rmse < (min(rmse) + sd(rmse))) %>%
      mutate(test = k_predictor + k_response) %>%
  arrange(test)

cv_dat %>%
  gather(type, metric, gam:cca) %>%
  mutate(rmse = map_dbl(metric, ~mean(.$rmse))) %>%
  select(-metric) %>%
    ggplot(aes(k_predictor, k_response, fill = rmse)) +
  geom_tile() +
  scale_fill_viridis_c(direction = -1) +
  facet_wrap(~type) +
  scale_x_continuous(breaks = 1:9, minor_breaks = NULL) +
  scale_y_continuous(breaks = 1:9, minor_breaks = NULL) +
  coord_equal() +
  theme_minimal()

cv_dat %>%
  gather(type, metric, gam:cca) %>%
  mutate(srmse = map_dbl(metric, ~sum(.$srmse > 0.7))) %>%
  select(-metric) %>%
  arrange(srmse) %>%
    filter(srmse < (min(srmse) + sd(srmse))) %>%
    mutate(test = k_predictor + k_response) %>%
  arrange(test)

cv_dat %>%
  gather(type, metric, gam:cca) %>%
  mutate(sspb = map_dbl(metric, ~mean(.$sspb))) %>%
  select(-metric) %>%
  arrange(sspb) %>%
    filter(sspb < (min(sspb) + sd(sspb))) %>%
  mutate(test = k_predictor + k_response) %>%

cv_dat$gam[[1]]
```

#### Domain averages
```{r}
wus_stats <- prism_dat %>%
  left_join(areas, by = c("x", "y")) %>%
  group_by(year) %>%
  #mutate(covered = if_else(SWE >= 3, area, 0)) %>% # 3mm threshold
    mutate(covered = pmin(1, SWE / 3) * area) %>% # linear decline below 3mm
  summarise(SWE = sum(SWE * area), # units in megaliters (?)
            SCA = sum(covered))

mountains_stats <- prism_dat %>%
  left_join(areas, by = c("x", "y")) %>%
  inner_join(mountains_mask, by = c("x", "y")) %>%
  group_by(year, name) %>%
 # mutate(covered = if_else(SWE >= 3, area, 0)) %>%
  mutate(covered = pmin(1, SWE / 3) * area) %>%
  summarise(SWE = sum(SWE * area), 
            SCA = sum(covered))
```

```{r}
cv_recons <- cv_dat %>%
  mutate(cv = map(cv, ~left_join(., areas, by = c("x", "y")))) %>%
  unnest(cv) %>%
 # mutate(covered = if_else(SWE >= 3, area, 0)) %>%
      mutate(covered = pmin(1, SWE / 3) * area) %>% # linear decline below 3mm
  group_by(k_predictor, k_response, year, type) %>%
  summarise(SWE = sum(SWE * area), 
            SCA = sum(covered)) %>%
  left_join(wus_stats, by = 'year', suffix = c('_recon', '_obs'))
gc()
range_recons <- cv_dat %>%
  mutate(cv = map(cv, ~left_join(., areas, by = c("x", "y")) %>% 
                    inner_join(mountains_mask, by = c("x", "y")))) %>%
  unnest(cv) %>%
 # mutate(covered = if_else(SWE >= 3, area, 0)) %>%
      mutate(covered = pmin(1, SWE / 3) * area) %>% # linear decline below 3mm
  group_by(k_predictor, k_response, name, year) %>%
  summarise(SWE = sum(SWE * area),
            SCA = sum(covered)) %>%
  left_join(mountains_stats, by = c('year', 'name'), suffix = c('_recon', '_obs'))
gc()
```


```{r}
rmse_dat <- cv_recons %>%
      summarise(SWE_rmse = sqrt(mean((SWE_obs - SWE_recon) ^ 2)),
                SCA_rmse = sqrt(mean((SCA_obs - SCA_recon) ^ 2)))
```

The cross validation routine selects `r rmse_dat[[1, "k_response"]]` PRISM PCs and  `r rmse_dat[[1, "k_predictor"]]` CERA PCs 
```{r}
arrange(rmse_dat, SWE_rmse)
arrange(rmse_dat, SCA_rmse)

rmse_dat %>%
  ggplot(aes(k_predictor, k_response, fill = SWE_rmse)) +
  geom_tile() +
  scale_fill_viridis_c(direction = -1) +
  scale_x_continuous(breaks = 1:9, minor_breaks = NULL) +
  scale_y_continuous(breaks = 1:9, minor_breaks = NULL) +
  coord_equal() +
  theme_minimal()

rmse_dat %>%
ggplot(aes(k_predictor, k_response, fill = SCA_rmse)) +
  geom_tile() +
  scale_fill_viridis_c(direction = -1) +
  scale_x_continuous(breaks = 1:9, minor_breaks = NULL) +
  scale_y_continuous(breaks = 1:9, minor_breaks = NULL) +
  coord_equal() +
  theme_minimal()

(rmse_dat$SWE_rmse %>% sd )+ 17704851
```


```{r, fig.asp = .35}
t1 <- cv_recons %>% filter(k_response == 6, k_predictor == 8) %>%
  gather(type, value, SWE_recon:SCA_obs) %>%
  separate(type, c('var', 'type')) %>%
  mutate(name = 'Western US')

ggplot(t1, aes(year)) +
  geom_line(aes(y = value, color = type)) +
  scale_color_manual(values = c('black', 'red')) +
    facet_wrap(~var, scales = 'free_y') +
  theme_bw()
```


```{r, fig.width = 8}
g <- range_recons %>%
  filter(k_response == 4, k_predictor == 4) %>%
  select(-SCA_recon, -SCA_obs) %>%
  rename(recon = SWE_recon, obs = SWE_obs) %>%
  gather(type, SWE, recon:obs) %>%
  ggplot(aes(year)) +
  geom_line(aes(y = SWE, color = type)) +
  scale_color_manual(values = c('black', 'red')) +
  facet_wrap(~name, scales = 'free_y', ncol = 3) +
  theme_bw()

h <- range_recons %>%
  filter(k_response == 7, k_predictor == 9) %>%
  select(-SWE_recon, -SWE_obs) %>%
  rename(recon = SCA_recon, obs = SCA_obs) %>%
  gather(type, SCA, recon:obs) %>%
  ggplot(aes(year)) +
  geom_line(aes(y = SCA, color = type)) +
  scale_color_manual(values = c('black', 'red')) +
    facet_wrap(~name, scales = 'free_y', ncol = 3) +
  theme_bw()

g / h + plot_annotation(tag_levels = 'A')
```

```{r, echo = FALSE, eval = FALSE}
# trying to combine above?
t2 <- range_recons %>%
  filter(k_response == 4, k_predictor == 4) %>%
   gather(type, value, SWE_recon:SCA_obs) %>%
  separate(type, c('var', 'type'))

bind_rows(t1, t2) %>%
   ggplot(aes(year)) +
  geom_line(aes(y = value, color = type)) +
  scale_color_manual(values = c('black', 'red')) +
    facet_grid(var~name, scales = 'free_y') +
  theme_bw()
```

#### Local error
```{r}
scores <- cv_dat %>%
  mutate(errors = map(cv, ~get_errors(.) %>% group_by(x, y) %>% filter(median(SWE_recon) > 3 & median(SWE_obs) > 3) %>% # or should it only befiltering by obs?
  ungroup()),
        score_map = map(errors, ~group_by(., x,y) %>% get_scores),
         score_ts = map(errors, ~group_by(., year) %>% get_scores)) %>%
  select(-cv, -errors)
```

time

```{r}
scores %>% mutate(k = interaction(k_response, k_predictor)) %>%
  dplyr::select(k, score_ts) %>%
  unnest(score_ts) %>%
  dplyr::select(k, mdsa, year) %>%
  ggplot(aes(year, mdsa)) +
  geom_line(aes(color = k)) +
  scale_color_viridis_d() +
  theme_bw()
```

```{r}
plot_fun1 <- function(x, name) {
  ggplot(x, aes(k_predictor, k_response, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(direction = -1, name = name) +
  coord_equal() +
      scale_x_continuous(breaks = 1:9, minor_breaks = NULL) +
  scale_y_continuous(breaks = 1:9, minor_breaks = NULL) +
  theme_minimal()
}
tmp1 <- scores %>%
  select(k_response, k_predictor, score_ts) %>%
  unnest(score_ts) %>%
  group_by(k_response, k_predictor) %>%
  summarise_at(vars(me:rmsle), mean) %>%
  gather(metric, value, me:rmsle) %>%
group_by(metric) %>%
  nest %>%
  mutate(plots = map2(data, metric, plot_fun1))

tmp1$plots
```

space 

```{r}
tmp2 <- scores %>%
  select(k_response, k_predictor, score_map) %>%
  unnest(score_map) %>%
  select(k_response:y, rmse, mdsa, srmse, rmsle) %>%
  group_by(k_response, k_predictor) %>%
  summarise(rmse = mean(rmse), mdsa = mean(mdsa), srmse_avg = mean(srmse), rmsle = mean(rmsle), srmse7 = sum(srmse>.7, na.rm = TRUE)) %>%
  gather(metric, value, rmse:srmse7) %>%
group_by(metric) %>%
  nest %>%
  mutate(plots = map2(data, metric, plot_fun1))

tmp2$plots
```

# Paleo



## older stuff

```{r}
t7 <- prism_dat %>%
  filter(year <= 2010) %>%
  spread(year, SWE) %>%
  rasterFromXYZ()

eot_rmse <- (predict(telecon, eot1) + mean(t7) - t7)^2 %>%
  cellStats(stat = 'mean') %>%
  sqrt()

cca_rmse <- tele_dat %>%
  mutate(swe_pred = tele_pred * swe_sd + swe_mean,
         error = swe_pred - SWE) %>% 
  group_by(year) %>%
  summarise(rmse = sqrt(mean(error ^ 2))) 

gam_rmse <- gam_recon %>%
  left_join(prism_dat, by = c('x', 'y', 'year')) %>%
  mutate(error = SWE.x - SWE.y) %>%
  group_by(year) %>%
  summarise(rmse = sqrt(mean(error ^ 2))) 

left_join(gam_rmse, cca_rmse, by = 'year') %>%
  mutate(eot_rmse = eot_rmse) %>%
  gather(method, rmse, rmse.x:eot_rmse) %>%
  ggplot(aes(year, rmse)) +
  geom_line(aes(color = method))
```

So the gam is in part effective here because it resolves recent snow droughts ... I wonder if this is a global warming trend?
```{r}
tele_dat %>%
  filter(year == 2005 ) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = anomaly)) +
  scale_fill_distiller(palette = 'RdBu', limits = c(-1100, 1100)) +
  facet_wrap(~year) +
  coord_quickmap()
```


```{r}
test2 <- tele.preds %>%
as_tibble() %>%
  bind_cols(t3, .) %>%
  rasterFromXYZ() 

plot(((test2 * calc(prism, sd) + mean(prism)) - prism )[[24]])
```

```{r}
plot(mean(prism) - prism)
mean(prism) - prism

tele.preds %>%
  as.tibble() %>%
  bind_cols(t3, .)
```
