---
title: "Downscaling snowpack variability in western North America"
subtitle: "Models and Observations"
author: "Nick Gauthier"
date: "Last knit on: `r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
link-citations: yes
---

Downscaling snow water equivalent 

# Setup

## Packages

Import the packages required for this analysis.

```{r setup}
# analysis
library(raster) # processing raster data
library(tidyverse) # data manipulation and visualization
library(mgcv) # flexible nonlinear regression models
library(remote)

# plotting
library(sf) # shapefiles and plotting
library(gganimate) # animated gifs
library(scico) # color palettes

# this package
devtools::load_all()
```

```{r}
load('../data.Rdata')
```


```{r, echo = FALSE}
# download state boundary files for cropping and plotting
state_names <- c('arizona', 'new mexico', 'colorado', 
                 'california', 'utah', 'nevada', 
                 'oregon', 'washington', 'idaho',
                 'wyoming', 'montana')

states_wus <- maps::map('state', regions = state_names, 
                    fill = TRUE, plot = FALSE) %>% st_as_sf
```

# Observed Modes of Snowpack Variability

Run the principal components analysis.
```{r}
prism_pcs <- calc_pcs(prism_dat, scale = FALSE)
```

Get the eigenvalues, and run some truncation tests.
```{r}
prism_eigs <- get_eigs(prism_pcs)
nFactors::nScree(prism_eigs$eigenvalues)
```

```{r, echo = FALSE}
plot_scree(prism_eigs, c(6, 14))
```
Lets start by truncating at 6. Extract the pc amplitude time series and EOF spatial patterns for the leading 6 modes.
```{r}
prism_patterns <- get_patterns(prism_pcs, prism_dat, prism_eigs, k = 6)
```

```{r, echo = FALSE}
plot_amps(prism_patterns)
```

```{r echo = FALSE}
plot_eof(prism_patterns)
```


# Simulated Modes

```{r}
cera_pca <- calc_pcs(cera_dat, scale = FALSE)
```



```{r, echo = FALSE}
cera_eigs <- get_eigs(cera_pca, cera)
```

```{r, echo = FALSE}


cera_eigs$eigenvalues %>%
  nScree() %>%
  plotnScree()
```


```{r plot_variance_obs, echo = FALSE}
plot_scree(cera_eigs, 7)
```


## EOFs

```{r calc_eofs}
prism_eofs <- get_eofs(prism_dat, prism_pca, prism_eigs, 7)
cera_eofs <- get_eofs(cera_dat, cera_pca, cera_eigs, 7)
```


```{r}
plot_eof(prism_eofs)
plot_eof(cera_eofs)
```

```{r}
pcs_prism <- get_pcs(prism_pca, 7)
pcs_cera <- get_pcs(cera_pca, 7)
```




```{r, echo = FALSE}
pcs_prism %>% 
  mutate(PC = paste0('PC', PC)) %>% 
ggplot(aes(year, amplitude)) +
  geom_line() +
  facet_wrap(~PC)

pcs_cera %>%
  mutate(PC = paste0('PC', PC)) %>% 
ggplot(aes(year, amplitude)) +
  geom_line() +
  facet_wrap(~PC)
```

```{r}
cera_pat <- left_join(group_nest(cera_eofs, EOF, .key = 'patterns'),
          group_nest(pcs_cera, PC, .key = 'amplitudes'),
          by = c('EOF' = 'PC'))

prism_pat <- left_join(group_nest(prism_eofs, EOF, .key = 'patterns'),
          group_nest(pcs_prism, PC, .key = 'amplitudes'),
          by = c('EOF' = 'PC'))
```


# Coupled Pattern Analysis

First look at just the error from the dimensionality reduction  with the PCs. This is basically the error in the pc targets, so is the minimum error we'd get even from the best downscaling

```{r}
## so this part is key, we are assuming the mean/sd are fixed and just using the values from the observations -- is there a better way?
prism_scaler <- prism_dat %>%
  spread(year, SWE) %>%
  dplyr::select(x, y) %>%
  mutate(avg = prism_pca$center, sdev = prism_pca$scale)

prism_recon <- pcs_prism %>%
  spread(year, amplitude) %>%
  left_join(prism_eofs, ., by = c('EOF' = 'PC')) %>%
  gather(year, amplitude, `1982`:`2017`) %>% ## change dates for cera
  mutate(SWE = value * amplitude,
         year = as.numeric(year)) %>%
  group_by(x, y, year) %>%
  summarise(anomaly = sum(SWE)) %>%
  left_join(prism_scaler) %>%
  mutate(SWE =  anomaly + avg)
```

```{r}
prism_recon %>%
  left_join(prism_dat, by = c('x', 'y', 'year')) %>%
  mutate(error = SWE.x - SWE.y) %>%
filter(year <= 1991) %>% filter(error < -500)
ggplot() +
  geom_raster(aes(x, y, fill = error)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  facet_wrap(~year) +
  #scale_fill_scico(palette = 'vik', direction = -1, limits = c(-1000, 1000)) +
  ggtitle('Error from dimensionality reduction', 'PRISM')
```

## GAM

```{r}
predictors <- pcs_cera %>%
  filter(year >= 1982) %>% 
    mutate(PC = paste0('PC', PC)) %>% 
  spread(PC, amplitude) %>%
  dplyr::select(year, PC1:PC7)

gam_mod <- pcs_prism %>%
  filter(year <= 2010) %>%
    mutate(PC = paste0('PC', PC)) %>% 
  left_join(predictors, by = 'year') %>%
  group_by(PC) %>%
  nest() %>%
  mutate(mod = map(data, ~gam(amplitude ~ 
                                s(PC1, k = 4) + 
                                s(PC2, k =5) + 
                                s(PC3, k = 4) + 
                                s(PC4, k = 4) + 
                                s(PC5, k = 4) + 
                                s(PC6, k = 4) + 
                                s(PC7, k = 4), 
                              data = ., 
                              method = 'REML', 
                              select = TRUE))) %>%
  mutate(r2 = map_dbl(mod, ~summary(.)$r.sq)) %>%
  mutate(data = map2(mod, data, ~mutate(.y, pred = predict(.x, .y)))) %>%
  dplyr::select(PC, data) %>%
  unnest(cols = c(data)) %>%
  dplyr::select(PC, amplitude, year, pred) %>%
  ungroup()
```

```{r}
  ggplot(gam_mod, aes(year)) +
  geom_line(aes(y = amplitude)) +
  geom_line(aes(y = pred), color = 'red') +
  facet_wrap(~PC)
```

Looks like pc 6 in the prism data may represent a trend ... r2 of .36 without year term, .89 with it!
this seems likely, year is selected out of the model for most of the other pcs, but shows a strong negative correlation in pc 6
```{r}
prism_eofs %>% filter(EOF == 6) %>% plot_eof()
```


```{r}
gam_recon <- gam_mod %>%
  dplyr::select(-amplitude) %>%
  spread(year, pred) %>% 
  mutate(PC = as.character(parse_number(PC))) %>%
  left_join(prism_eofs, ., by = c('EOF' = 'PC')) %>%
  gather(year, amplitude, `1982`:`2010`) %>% ## change dates for cera
  mutate(SWE = value * amplitude,
         year = as.numeric(year)) %>%
  group_by(x, y, year) %>%
  summarise(anomaly = sum(SWE)) %>%
  left_join(prism_scaler) %>%
  mutate(SWE =  anomaly + avg)
```


```{r}
gam_recon %>%
filter(year <= 1987) %>% 
  mutate(SWE = if_else(SWE < 0, 0, SWE)) %>%
ggplot() +
  geom_raster(aes(x, y, fill = anomaly)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  facet_wrap(~year) +
  scale_fill_scico(palette = 'vik', direction = -1, limits = c(-1300, 1300)) +
  ggtitle('Observed mean April 1st SWE, 1982-2017', 'PRISM')

prism_dat %>%
    group_by(x, y) %>%
  mutate(SWE_mean = mean(SWE),
         anomaly = SWE - SWE_mean) %>% 
filter(year <= 1987) %>%
ggplot() +
  geom_raster(aes(x, y, fill = anomaly)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  facet_wrap(~year) +
  scale_fill_scico(palette = 'vik', direction = -1, limits = c(-1300, 1300)) +
  ggtitle('Observed mean April 1st SWE, 1982-2017', 'PRISM')
```


```{r}
gam_recon %>%
  left_join(prism_dat, by = c('x', 'y', 'year')) %>%
  mutate(error = SWE.x - SWE.y) %>%
filter(year <= 1987) %>%
ggplot() +
  geom_raster(aes(x, y, fill = error)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  facet_wrap(~year) +
  scale_fill_scico(palette = 'vik', direction = -1, limits = c(-450, 450)) +
  ggtitle('Observed mean April 1st SWE, 1982-2017', 'PRISM')
```


## CCA

```{r}
t1 <- pcs_prism %>%
  filter(year <= 2010) %>%
  spread(PC, amplitude) %>%
  dplyr::select(-year) %>%
  as.matrix() %>%
  scale()

t2 <- pcs_cera %>%
  filter(year >= 1982) %>% 
    spread(PC, amplitude) %>%
  dplyr::select(-year) %>%
  as.matrix() %>%
  scale()

can_cor<- cancor(t1, t2, xcenter = FALSE, ycenter = FALSE) # pcs already centered
```

```{r}
can_cor$cor
```

```{r}
cera_can <- cera_eofs %>%
  dplyr::select(-value) %>%  # use normlaized or unnormalized?
  spread(EOF, weight) %>%
  dplyr::select(-x, -y) %>%
  as.matrix() %>%
  `%*%`(can_cor$ycoef) %>%
    `colnames<-`(paste0('PC', 1:7)) %>%
  as_tibble()
  
prism_can <- prism_eofs %>%
  dplyr::select(-value) %>%
  spread(EOF, weight) %>%
  dplyr::select(-x, -y) %>%
  as.matrix() %>%
  `%*%`(can_cor$xcoef) %>%
  `colnames<-`(paste0('PC', 1:7)) %>%
  as_tibble()
```

```{r}
cera_dat %>%
  spread(year, SWE) %>%
  mutate(column = 1:n()) %>%
  dplyr::select(x, y, column) %>%
  bind_cols(cera_can) %>%
  gather(pattern, value, PC1:PC7) %>% 
  ggplot() +
  geom_raster(aes(x, y, fill = value)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~pattern) +
      scale_fill_scico(palette = 'broc', direction = -1, limits = c(-1, 1) * 8) +
  theme_void()+
  coord_sf() +
  ggtitle('Reanalysis CCA')

prism_dat %>%
  spread(year, SWE) %>%
  mutate(column = 1:n()) %>%
  dplyr::select(x, y, column) %>%
  bind_cols(prism_can) %>%
  gather(pattern, value, PC1:PC7) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = value)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~pattern) +
      scale_fill_scico(palette = 'broc', direction = -1, limits = c(-1, 1) * 80) +
  theme_void()+
  coord_sf() +
  ggtitle('Observed CCA')
```

```{r}
library(patchwork)
plot_cancor <- function(x1, x2, y1, y2){
  p1 <- x1 %>%
  spread(year, SWE) %>%
  mutate(column = 1:n()) %>%
  dplyr::select(x, y, column) %>%
  bind_cols(x2) %>%
  gather(pattern, value, PC1:PC7) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = value)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~pattern) +
  scale_fill_distiller(palette = 'RdBu') +
  theme_void()+
  #coord_quickmap() +
  ggtitle('Reanalysis CCA') +
    theme(legend.position = 'bottom')

p2 <- y1 %>%
  spread(year, SWE) %>%
  mutate(column = 1:n()) %>%
  dplyr::select(x, y, column) %>%
  bind_cols(y2) %>%
  gather(pattern, value, PC1:PC7) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = value)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  facet_wrap(~pattern) +
  scale_fill_distiller(palette = 'RdBu') +
  theme_void()+
  #coord_quickmap() +
  ggtitle('Observed CCA') +
  theme(legend.position = 'bottom')

p1 + p2
}

plot_cancor(cera_dat, cera_can, prism_dat, prism_can)
```


```{r}
library(telefit)

t1 <- prism_dat %>%
  filter(year <= 2010) %>%
  snow_only() %>%
  spread(year, SWE) %>%
  dplyr::select(-x, -y) %>%
  as.matrix()

t2 <- cera_dat %>%
  filter(year >= 1982) %>% 
  snow_only() %>%
  spread(year, SWE) %>%
  dplyr::select(-x, -y) %>%
  as.matrix()


t3 <- prism_dat %>%
  filter(year <= 2010) %>%
  snow_only() %>%
  spread(year, SWE) %>%
  dplyr::select(x, y)

tele.preds <- cca.predict(t2, t1, t2, k.x = 7, k.y = 4)
1 - var(as.numeric(tele.preds-t1)) / var(as.numeric(t1))
```

it looks like for some reason the cca.predict function isn't rescaling correctly? or its just meant to work with standardized anomalies?

```{r}
prism_scaler <- prism_dat %>%
  group_by(x,y) %>%
  summarise(avg = mean(SWE), sdev = sd(SWE, na.rm = TRUE))
```

```{r}
prism_dat %>%
  filter(year <= 2010) %>%
  snow_only() %>%
  spread(year, SWE) %>%
  dplyr::select(x, y) %>%
  cbind(tele.preds) %>%
  gather(year, SWE, `1982`:`2010`) %>%
  left_join(prism_scaler) %>%
  mutate(SWE = SWE * sdev + avg) %>%
  filter(year <=1984) %>%
  ggplot(aes(x,y)) +
  geom_raster(aes(fill = SWE)) +
  facet_wrap(~year) +
  coord_quickmap()
```


```{r}
cca.predict
```


## EOTs

```{r}
eot1 <- cera_dat %>%
  filter(year >= 1982) %>%
  spread(year, SWE) %>%
  rasterFromXYZ() %>%
  anomalize() %>%
  denoise(expl.var = .85, weighted = TRUE)

eot2 <- prism_dat %>%
  filter(year <= 2010) %>%
  spread(year, SWE) %>%
  rasterFromXYZ() %>%
  anomalize() %>% 
  denoise(expl.var = .85, weighted = TRUE)
```

```{r}
telecon <- eot(eot1, eot2, n = 6)
```


```{r}
walk(1:6, ~plot(telecon, y = ., show.bp = TRUE))
```

```{r}
predict(telecon, eot1) %>% .[[1]] %>% `+`(mean(prism)) %>% plot
```

# Validation


```{r}
anomaly_dat <- prism_dat %>%
  filter( year <= 2010) %>% # important to make sure everything's calculated on the same time scale, may be best to do this further up?
  group_by(x, y) %>%
  mutate(swe_mean = mean(SWE),
         swe_sd = sd(SWE, na.rm = TRUE),
         anomaly = SWE - swe_mean) %>%
  ungroup()

tele_dat <- tele.preds %>%
  as_tibble() %>%
  bind_cols(t3, .) %>%
  gather(year, tele_pred, `1982`:`2010`) %>%
  mutate(year = as.numeric(year)) %>%
  left_join(anomaly_dat)  %>%
  mutate(swe_pred = tele_pred * swe_sd + swe_mean,
         res = swe_pred - SWE) 
```

```{r, echo = FALSE}
tele_dat %>%
  filter(year < 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = tele_pred)) +
  scale_fill_viridis_c() +
  facet_wrap(~year) +
  coord_quickmap()

tele_dat %>%
  filter(year < 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = anomaly)) +
  scale_fill_distiller(palette = 'RdBu') +
  facet_wrap(~year) +
  coord_quickmap()
```

```{r, echo = FALSE}
tele_dat %>%
  filter(year == 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = swe_mean)) +
  scale_fill_viridis_c() +
  facet_wrap(~year) +
  coord_quickmap()

tele_dat %>%
  filter(year == 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = swe_sd)) +
  scale_fill_viridis_c() +
  facet_wrap(~year) +
  coord_quickmap()
```

```{r}
plot(mean(prism));plot(calc(prism, sd))
```

```{r}
tele_dat %>%
  mutate(test = tele_pred * swe_sd + swe_mean) %>%
    filter(year == 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = test)) +
  scale_fill_viridis_c() +
  facet_wrap(~year) +
  coord_quickmap()

tele_dat %>%
  mutate(test = tele_pred * swe_sd + swe_mean) %>%
    filter(year == 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = SWE)) +
  scale_fill_viridis_c() +
  facet_wrap(~year) +
  coord_quickmap()
```

```{r}
tele_dat %>%
  mutate(swe_pred = tele_pred * swe_sd + swe_mean,
         error = swe_pred - SWE) %>%
      filter(year < 1985) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = error)) +
  scale_fill_distiller(palette = 'RdBu') +
  facet_wrap(~year) +
  coord_quickmap()
```

```{r}
t7 <- prism_dat %>%
  filter(year <= 2010) %>%
  spread(year, SWE) %>%
  rasterFromXYZ()

eot_rmse <- (predict(telecon, eot1) + mean(t7) - t7)^2 %>%
  cellStats(stat = 'mean') %>%
  sqrt()

cca_rmse <- tele_dat %>%
  mutate(swe_pred = tele_pred * swe_sd + swe_mean,
         error = swe_pred - SWE) %>% 
  group_by(year) %>%
  summarise(rmse = sqrt(mean(error ^ 2))) 

gam_rmse <- gam_recon %>%
  left_join(prism_dat, by = c('x', 'y', 'year')) %>%
  mutate(error = SWE.x - SWE.y) %>%
  group_by(year) %>%
  summarise(rmse = sqrt(mean(error ^ 2))) 

left_join(gam_rmse, cca_rmse, by = 'year') %>%
  mutate(eot_rmse = eot_rmse) %>%
  gather(method, rmse, rmse.x:eot_rmse) %>%
  ggplot(aes(year, rmse)) +
  geom_line(aes(color = method))
```

So the gam is in part effective here because it resolves recent snow droughts ... I wonder if this is a global warming trend?
```{r}
tele_dat %>%
  filter(year == 2005 ) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = anomaly)) +
  scale_fill_distiller(palette = 'RdBu', limits = c(-1100, 1100)) +
  facet_wrap(~year) +
  coord_quickmap()
```


```{r}
test2 <- tele.preds %>%
as_tibble() %>%
  bind_cols(t3, .) %>%
  rasterFromXYZ() 

plot(((test2 * calc(prism, sd) + mean(prism)) - prism )[[24]])
```

```{r}
plot(mean(prism) - prism)
mean(prism) - prism

tele.preds %>%
  as.tibble() %>%
  bind_cols(t3, .)
```


```{r, echo = FALSE}
knitr::knint_exit()
```


Some aggregate statistics. Is this the right way to aggregate?

```{r}
obs_agg <- prism_dat %>%
  filter(SWE > 0) %>%
  group_by(year) %>%
  summarise(SWE = sum(SWE * area), # units in megaliters (?)
            area = sum(area))
```

How has total SWE varied over the observational period?
```{r, echo = FALSE, fig.fullwidth = TRUE}
ggplot(obs_agg, aes(year, SWE)) +
  #geom_vline(xintercept = 2015, color = 'red', linetype = 2) +
  geom_col(aes(fill = SWE)) +
  scale_fill_viridis_c(guide = FALSE) +
  theme_minimal() +
  ggtitle('Total SWE over western North America')
```

```{r, echo = FALSE, fig.fullwidth = TRUE}
ggplot(obs_agg, aes(year, SWE)) +
  #geom_vline(xintercept = 2015, color = 'red', linetype = 2) +
  geom_point(aes(color = SWE)) +
  scale_color_viridis_c(guide = FALSE) +
  geom_smooth(method = 'lm') +
  theme_minimal() +
  ggtitle('Total SWE over western North America')
```

What about total area covered?
```{r, echo = FALSE, fig.fullwidth = TRUE}
ggplot(obs_agg, aes(year, area)) +
  geom_col(aes(fill = area)) +
  scale_fill_viridis_c(guide = FALSE) +
  theme_minimal() +
  ggtitle('Total snow area over western North America')
```

```{r, echo = FALSE}
ggplot(obs_agg, aes(area, SWE)) +
  geom_point(aes(color = year), size = 2.5) +
  scale_color_viridis_c() +
  geom_smooth(method = 'lm') +
  theme_minimal() +
  labs(x = 'Area (km2)', title = 'Relationship between total SWE and snow extent')
```


# Visualization



What's the climatic mean April 1st SWE for the observations and simulations?
```{r, echo = FALSE}
prism_dat %>%
  group_by(x, y) %>%
  summarise(SWE = mean(SWE))%>%
  ggplot() +
  geom_raster(aes(x, y, fill = SWE)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  scale_fill_scico(palette = 'davos', direction = -1) +
  ggtitle('Observed mean March SWE, 1982-2017', 'PRISM')
```

```{r}
cera_dat %>%
  group_by(x, y) %>%
  summarise(SWE = mean(SWE))%>%
  ggplot() +
  geom_raster(aes(x, y, fill = SWE)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  scale_fill_scico(palette = 'davos', direction = -1) +
  ggtitle('Reanalysis mean March SWE, 1900-2010', 'PRISM')
```


To what extent is this mistmatch due to bias or scale effects? Test by resampling the observations to the CERA grid.

```{r, echo = FALSE}
mean_swe_cera <- cera_dat %>%
  filter(year >= 1982) %>%
  group_by(x, y) %>%
  summarise(sim = mean(SWE))

prism_resamp <- prism %>%
  subset(1:29) %>%
  mean %>%
  resample(cera) %>%
  as.data.frame(xy = TRUE, na.rm = TRUE) %>%
  rename(obs = layer)

left_join(prism_resamp, mean_swe_cera, by = c('x', 'y')) %>%
  gather(type, SWE, 3:4) %>%
ggplot() +
  geom_raster(aes(x, y, fill = SWE)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  facet_wrap(~type) +
    scale_fill_scico(palette = 'davos', direction = -1) +
  ggtitle('Observed vs Simulated mean March SWE', '1982-2010')
```


Look at how SWE has varied over time and space.

```{r animation, echo = FALSE, warning=FALSE,cache = TRUE}
prism_dat %>%
  ggplot() +
  geom_raster(aes(x, y, fill = SWE)) +
    scale_fill_scico(palette = 'davos', direction = -1) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  ggtitle("Snow Water Equivalent", "March, {closest_state}") +
  transition_states(year)
```

```{r cera_animation, echo = FALSE, warning=FALSE,cache = TRUE}
cera_dat %>%
  ggplot() +
  geom_raster(aes(x, y, fill = SWE)) +
    scale_fill_scico(palette = 'davos', direction = -1) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  theme_void() +
  ggtitle("Snow Water Equivalent", "March, {closest_state}") +
  transition_states(year)
```

```{r, echo = FALSE}
prism_dat %>%
  group_by(x, y) %>%
  mutate(anomaly = SWE - mean(SWE)) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = anomaly)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  scale_fill_scico(palette = 'broc',
                   direction = -1,
                   limits = c(-1300, 1300)) +
  transition_states(year) +
  theme_void()

cera_dat %>%
  group_by(x, y) %>%
  mutate(anomaly = SWE - mean(SWE)) %>%
  ggplot() +
  geom_raster(aes(x, y, fill = anomaly)) +
  geom_sf(data = states_wus, fill = NA, color = 'black') +
  scale_fill_scico(palette = 'broc',
                   direction = -1,
                   limits = c(-130, 130)) +
  transition_states(year) +
  theme_void()
```



So there's a clear bias here. As we see below, the mean of the observations is well higher than it should be. Looks like the simulation underpredicts snow -- likely because of all the small-scale topographic influences that CESM can't resolve.

