---
title: "data_prep"
author: "Nick Gauthier"
date: "2/26/2020"
output: html_document
---


```{r setup}
# this package
devtools::load_all()
library(raster) # processing raster data
library(tidyverse) # data manipulation and visualization
```

## Geographic Data

Define a study area to constrain all computations.
```{r}
bbox <- extent(c(-125, -102, 31, 49))
```

Import the snow observation data from https://nsidc.org/data/nsidc-0719.^[What's up with this warning message? long_name=CRS definition
spatial_ref=GEOGCS["NAD83",DATUM["North_American_Datum_1983",SPHEROID["GRS 1980",6378137,298.257222101,AUTHORITY["EPSG","7019"]],AUTHORITY["EPSG","6269"]],PRIMEM["Greenwich",0,AUTHORITY["EPSG","8901"]],UNIT["degree",0.01745329251994328,AUTHORITY["EPSG","9122"]],AUTHORITY["EPSG","4269"]]
GeoTransform=-125.0208 0.04166662697178698 0 49.9375 0 -0.04166662697178698]. 

```{r}
preprocess <- function(x, var, bbox, flip = FALSE, regrid = FALSE, daily = FALSE) {
  maps <- brick(x, varname = var)

  indices <- getZ(maps) %>% # find the index for April 1st
    str_detect('-03-')%>%
    which()

  raster::subset(maps, indices) %>%
    {if(flip) rotate(.) else .} %>%
    crop(bbox) %>%
    {if(daily) mean(.) else .} %>%
    {if(regrid) aggregate(., fact = 2) else .} # resample to lower res to speed up analysis
}
```

```{r message=FALSE,warning=FALSE}
prism <- list.files('../data', pattern = 'SWE_Depth', full.names = TRUE) %>%
  map(preprocess, var = 'SWE', bbox = bbox, regrid = TRUE, daily = TRUE) %>%
  brick() %>%
  setNames(1982:2017)
```


```{r, warning = FALSE}
# this averages over the full ensemble
cera <- map(1:10,
            ~ (brick('../data/CERA-20C_snow.nc', varname = 'rsn', level = .) *
                 brick('../data/CERA-20C_snow.nc', varname = 'sd', level = .))) %>%
  reduce(`+`) %>%
  `/`(10) %>%
  crop(bbox)
```


```{r}
livneh <- preprocess('~/Downloads/LIVNEH/swe.mon.mean.nc', var = 'swe', bbox = bbox, flip = TRUE, regrid = TRUE)
```

## Preprocessing
```{r}
snow_only <- function(dat){
  dat %>%
    group_by(x, y) %>%
    mutate(test = all(near(SWE, 0) | is.na(SWE))) %>%
    filter(test == FALSE) %>%
    ungroup() %>%
    dplyr::select(-test)
}
```

Turn the raster bricks into data frames and join.
```{r}
prism_dat <- prism %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  dplyr::select(-layer) %>%
  snow_only()
```

```{r}
cera_dat <- cera %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  dplyr::select(-layer) %>%
  snow_only()
```

```{r}
livneh_dat <- livneh %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  dplyr::select(-layer) %>%
  snow_only()
```

```{r}
save(prism, prism_dat, cera, cera_dat, livneh, livneh_dat, file = '../data.Rdata')
```
