---
title: "data_prep"
author: "Nick Gauthier"
date: "2/26/2020"
output: html_document
---


```{r setup}
library(raster) # processing raster data
library(tidyverse) # data manipulation and visualization
library(sf)

# download state boundary files for cropping and plotting
state_names <- c('arizona', 'new mexico', 'colorado', 
                 'california', 'utah', 'nevada', 
                 'oregon', 'washington', 'idaho',
                 'wyoming', 'montana')

states_wus <- maps::map('state', regions = state_names, 
                    fill = TRUE, plot = FALSE) %>% 
  st_as_sf()
```

## Geographic Data

Define a study area to constrain all computations.
```{r}
bbox <- extent(c(-125, -102, 31, 49))
```

Import the snow observation data from https://nsidc.org/data/nsidc-0719.^[What's up with this warning message? long_name=CRS definition
spatial_ref=GEOGCS["NAD83",DATUM["North_American_Datum_1983",SPHEROID["GRS 1980",6378137,298.257222101,AUTHORITY["EPSG","7019"]],AUTHORITY["EPSG","6269"]],PRIMEM["Greenwich",0,AUTHORITY["EPSG","8901"]],UNIT["degree",0.01745329251994328,AUTHORITY["EPSG","9122"]],AUTHORITY["EPSG","4269"]]
GeoTransform=-125.0208 0.04166662697178698 0 49.9375 0 -0.04166662697178698]. 

```{r}
preprocess <- function(x, var, bbox, flip = FALSE, regrid = FALSE, daily = FALSE) {
  maps <- brick(x, varname = var)

  indices <- getZ(maps) %>% # find the index for April 1st
    str_detect('-03-')%>%
    which()

  raster::subset(maps, indices) %>%
    {if(flip) rotate(.) else .} %>%
    crop(bbox) %>%
    {if(daily) mean(.) else .} %>% # if a file is one year of daily values
    {if(regrid) aggregate(., fact = 2) else .} # resample to lower res to speed up analysis
}
```

```{r message=FALSE,warning=FALSE}
prism <- list.files('../data', pattern = 'SWE_Depth', full.names = TRUE) %>%
  map(preprocess, var = 'SWE', bbox = bbox, regrid = TRUE, daily = TRUE) %>%
  brick() %>%
  mask(., all(near(., 0)), maskvalue = 1) %>% # mask pixels that never receive snow
  mask(states_wus) %>% # crop to state boundaries
  setNames(1982:2017)
```

```{r, warning = FALSE}
cera <- map(1:10, ~ (brick('../data/CERA-20C_snow.nc', varname = 'sd', level = .))) %>%
  # this averages over the full ensemble
  reduce(`+`) %>%
  `/`(10) %>%
  crop(bbox) %>%
  mask(., all(near(., 0)), maskvalue = 1) %>%
  mask(states_wus) %>%
  `*`(1000) # convert to mm
```

```{r warning=FALSE}
cesm_h2osno <- preprocess('../data/b.e11.BLMTRC5CN.f19_g16.001.clm2.h0.H2OSNO.085001-184912.nc',
                          var = 'H2OSNO',
                          flip = TRUE,
                          bbox = bbox)

cesm_h2osno_ext <- preprocess('../data/b.e11.BLMTRC5CN.f19_g16.001.clm2.h0.H2OSNO.185001-200512.nc',
                          var = 'H2OSNO',
                          flip = TRUE,
                          bbox = bbox)

cesm <- c(cesm_h2osno, cesm_h2osno_ext) %>%
  brick() %>%
  resample(cera) %>%
  mask(mean(cera))
```

```{r}
ccsm <- preprocess('../../snow-wna/data/snw_LImon_CCSM4_past1000_r1i1p1_085001-185012.nc',
                   var = 'snw',
                   flip = TRUE,
                   bbox = bbox) %>%
  resample(cera) %>%
  mask(mean(cera))
```

```{r}
plot(mean(prism));plot(mean(cera));plot(mean(cesm));plot(mean(ccsm))
```
supplemental data
```{r}
areas_prism <- area(prism) %>%
  as.data.frame(xy = TRUE, na.rm = TRUE) %>%
  rename(area = layer)

areas_cera <- area(cera) %>%
  as.data.frame(xy = TRUE, na.rm = TRUE) %>%
  rename(area = layer)

mountains <- read_sf('../data/ne_10m_geography_regions_polys.shp') %>%
  filter(name %in% c('SIERRA NEVADA', 'CASCADE RANGE', 'ROCKY MOUNTAINS')) %>%
  dplyr::select(name, geometry)

mountains_mask <- mountains %>% 
  mutate(rasters = split(., 1:3) %>% map(~mask(prism[[1]], .))) %>%
  st_drop_geometry() %>%
  mutate(rasters = map(rasters, as.data.frame, na.rm = TRUE, xy = TRUE)) %>%
  unnest(rasters) %>%
  dplyr::select(-X1982)
```



Turn the raster bricks into data frames and join.
```{r}
prism_dat <- prism %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  dplyr::select(-layer)
```

```{r}
cera_dat <- cera %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  dplyr::select(-layer)
```


still need to account for snow fraction of grid cell? and land fraction too?

```{r}
cesm_dat <- cesm %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = str_sub(layer, 2) %>% parse_number() %>% round) %>%
  rename(SWE = value) %>%
  select(-layer)
```

```{r}
ccsm_dat <- ccsm %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = str_sub(layer, 2) %>% parse_number() %>% round) %>%
  rename(SWE = value) %>%
  select(-layer)
```

```{r}
save(prism_dat, cera_dat, cesm_dat, ccsm_dat, areas_prism, areas_cera, states_wus, file = '../data.Rdata')
```

## Bilinear interpolation


```{r}
raster::resample(cera[[99]] / mean(cera[[82:110]]), prism) %>% plot
raster::resample((cera[[99]] / mean(cera[[82:110]])), prism) %>% plot
  

filter(year >= 1982) %>%
    group_by(x,y) %>%
  mutate(SWE = (SWE - mean(SWE))/sd(SWE)) %>%
  ungroup() %>%
  filter(year == 1999) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = SWE)) +
  coord_quickmap() +
  scale_fill_viridis_c(limits = c(-5,5))
```

```{r}
cera[cera < 0.001] <- 0

delta_mult <- raster::resample(cera / mean(cera[[82:110]]) , prism) * mean(prism)
delta_mult2 <- raster::resample(cera / mean(cera) , prism) * mean(prism)

delta_add <- raster::resample(cera - mean(cera[[82:110]]) , prism) + mean(prism)
```
