---
title: "data_prep"
author: "Nick Gauthier"
date: "2/26/2020"
output: html_document
---


```{r setup}
# analysis
library(raster) # processing raster data
library(tidyverse) # data manipulation and visualization
library(furrr) # parallel processing

# this package
devtools::load_all()
```


```{r}
plan(multisession)
```

## Geographic Data

Define a study area to constrain all computations.
```{r}
bbox <- extent(c(-125, -104, 33, 49))
```

Import the snow observation data from https://nsidc.org/data/nsidc-0719.^[What's up with this warning message? long_name=CRS definition
spatial_ref=GEOGCS["NAD83",DATUM["North_American_Datum_1983",SPHEROID["GRS 1980",6378137,298.257222101,AUTHORITY["EPSG","7019"]],AUTHORITY["EPSG","6269"]],PRIMEM["Greenwich",0,AUTHORITY["EPSG","8901"]],UNIT["degree",0.01745329251994328,AUTHORITY["EPSG","9122"]],AUTHORITY["EPSG","4269"]]
GeoTransform=-125.0208 0.04166662697178698 0 49.9375 0 -0.04166662697178698]. 

```{r message=FALSE,warning=FALSE}
# import SWE data

preprocess_prism <- function(x, var = 'SWE'){
  maps <- brick(x, varname = var)

  indices <- getZ(maps) %>% # find the indices for March
    str_detect('-03-') %>%
    which()

  raster::subset(maps, indices) %>%
    crop(bbox) %>%
    mean() %>%
    raster::aggregate(fact = 3) # resample to lower res to speed up analysis
}

prism <- list.files('../data', pattern = 'SWE_Depth', full.names = TRUE) %>%
  future_map(preprocess_prism, var = 'SWE') %>%
  brick() %>%
  setNames(1982:2017)
```


```{r, warning = FALSE}
# this averages over the full ensemble
cera <- map(1:10,
            ~ (brick('../data/CERA-20C_snow.nc', varname = 'rsn', level = .) *
                 brick('../data/CERA-20C_snow.nc', varname = 'sd', level = .))) %>%
  reduce(`+`) %>%
  `/`(10) %>%
  crop(bbox)
```

## Preprocessing

Turn all the rasters into data frames and join.
```{r}
prism_dat <- prism %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  select(-layer) %>%
  snow_only()
```

```{r}
cera_dat <- cera %>%
  as.data.frame(xy = TRUE, na.rm = TRUE, long = TRUE) %>%
  mutate(year = round(parse_number(layer))) %>%
  rename(SWE = value) %>%
  select(-layer) %>%
  snow_only()
```

```{r}
save(prism, prism_dat, cera, cera_dat, file = '../data.Rdata')
```
