---
title: "Statistical Downscaling with CCA"
format: html
vignette: >
  %\VignetteIndexEntry{Statistical Downscaling with CCA}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| message: false
library(tidyeof)
library(stars)
library(dplyr)
library(ggplot2)
```

## Overview

`tidyeof` supports statistical downscaling via Canonical Correlation Analysis (CCA). The workflow is:

1. Extract EOF patterns from both predictor (coarse) and response (fine) fields
2. Couple them with CCA to find correlated modes of co-variability
3. Predict fine-scale fields from new coarse-scale data

This vignette demonstrates the full downscaling pipeline.

## Setup: predictor and response data

For this vignette, we create a synthetic downscaling problem. The "fine" field is the bundled PRISM temperature data. The "coarse" field is derived by spatially aggregating it (simulating a coarse-resolution model).

```{r}
#| label: load-data
# Fine-resolution "observations"
fine <- system.file("testdata/prism_test.RDS", package = "tidyeof") |>
  readRDS()

# Coarse-resolution "model" -- aggregate to ~5x coarser grid
# In practice, this would be a GCM or reanalysis product
coarse <- fine |>
  st_warp(cellsize = 0.2, method = "average", use_gdal = TRUE, no_data_value = -99999) |>
  setNames(names(fine))
```

```{r}
#| label: show-grids
#| fig-width: 8
#| fig-height: 3
p1 <- ggplot() + geom_stars(data = slice(fine, "time", 1)) +
  scale_fill_viridis_c(na.value = NA) + coord_sf() + theme_void() + ggtitle("Fine (PRISM)")
p2 <- ggplot() + geom_stars(data = slice(coarse, "time", 1)) +
  scale_fill_viridis_c(na.value = NA) + coord_sf() + theme_void() + ggtitle("Coarse (aggregated)")
patchwork::wrap_plots(p1, p2)
```

## Step 1: Extract patterns

Extract EOF patterns from both fields. The number of modes `k` controls how much spatial detail is retained. More modes capture finer structure but risk overfitting in the CCA step.

```{r}
#| label: patterns
pred_pat <- patterns(coarse, k = 5)
resp_pat <- patterns(fine, k = 5)
```

```{r}
#| label: plot-patterns
#| fig-width: 10
#| fig-height: 4
plot(pred_pat, type = "eofs") + ggtitle("Predictor EOFs (coarse)")
```

```{r}
#| label: plot-resp-patterns
#| fig-width: 10
#| fig-height: 4
plot(resp_pat, type = "eofs") + ggtitle("Response EOFs (fine)")
```

## Step 2: Couple with CCA

`couple()` finds the linear combinations of predictor and response PCs that are maximally correlated. The `k` argument controls how many CCA modes to retain.

```{r}
#| label: couple
coupled <- couple(pred_pat, resp_pat, k = 3)
coupled
```

### Diagnostics

Inspect the canonical correlations -- these indicate the strength of the predictor-response relationship for each mode:

```{r}
#| label: correlations
get_canonical_correlations(coupled)
```

```{r}
#| label: plot-correlations
#| fig-width: 5
#| fig-height: 3
plot(coupled, type = "correlations")
```

Visualize canonical variates (the optimally-correlated time series from each side):

```{r}
#| label: canonical-variates
#| fig-width: 8
#| fig-height: 4
plot(coupled, type = "canonical", side = "both")
```

Canonical spatial patterns show what spatial structures in the predictor correspond to what structures in the response:

```{r}
#| label: canonical-patterns
#| fig-width: 8
#| fig-height: 6
plot(coupled, type = "canonical_patterns", side = "both")
```

The combined view shows everything together:

```{r}
#| label: combined-view
#| fig-width: 10
#| fig-height: 12
plot(coupled)
```

## Step 3: Predict

`predict()` takes new coarse-resolution data and produces a fine-resolution prediction. Internally it: projects the new data onto predictor EOFs, applies the CCA transform, multiplies by response EOFs, and adds the response climatology.

```{r}
#| label: predict
# Use the training data as a test (perfect predictor scenario)
predicted <- predict(coupled, coarse)
```

```{r}
#| label: compare
#| fig-width: 8
#| fig-height: 6
# Compare a single time step
t_idx <- 12
p1 <- ggplot() + geom_stars(data = slice(fine, "time", t_idx)) +
  scale_fill_viridis_c(na.value = NA) + coord_sf() + theme_void() +
  ggtitle("Observed (fine)")
p2 <- ggplot() + geom_stars(data = slice(predicted, "time", t_idx)) +
  scale_fill_viridis_c(na.value = NA) + coord_sf() + theme_void() +
  ggtitle("Predicted (downscaled)")
patchwork::wrap_plots(p1, p2)
```

### Amplitudes only

If you don't need the full spatial reconstruction (e.g., for diagnostics), set `reconstruct = FALSE` to get predicted response amplitudes directly:

```{r}
#| label: amps-only
pred_amps <- predict(coupled, coarse, reconstruct = FALSE)
head(pred_amps)
```

## Common patterns for multi-source downscaling

When downscaling across multiple data sources (e.g., multiple reanalyses or model runs), use `common_patterns()` to extract shared EOF patterns. This ensures all sources are projected into the same EOF space, making their amplitudes directly comparable.

```{r}
#| label: common-patterns
# Simulate two "sources" by splitting the time series
times <- st_get_dimension_values(coarse, "time")
source_a <- filter(coarse, time <= times[18])
source_b <- filter(coarse, time > times[18])

cpat <- common_patterns(list(a = source_a, b = source_b), k = 4, weight = TRUE)
cpat
```

Each source gets its own patterns object with source-specific amplitudes and climatology, but shared EOF spatial patterns:

```{r}
#| label: common-congruence
# The spatial patterns are identical; only amplitudes and climatology differ
cpat$a$eofs
cpat$b$eofs
```

You can then couple one source's patterns with fine-scale observations, and predict using another source's patterns via the `predictor_patterns` argument:

```{r}
#| label: cross-source
#| eval: false
# Train CCA using source A
coupled_a <- couple(cpat$a, resp_pat, k = 3)

# Predict using source B's data, with B's climatology
predict(coupled_a, source_b_new, predictor_patterns = cpat$b)
```

## Next steps

- See `vignette("cross-validation")` for tuning `k_pred`, `k_resp`, and `k_cca`
- The `reconstruct()` function can also be used standalone for EOF-based field reconstruction (see `vignette("eof-analysis")`)

## References

- Barnett, T.P. & Preisendorfer, R. (1987). Origins and levels of monthly and seasonal forecast skill for United States surface air temperatures determined by canonical correlation analysis. *Monthly Weather Review*, 115(9), 1825-1850.
- Bretherton, C.S., Smith, C., & Wallace, J.M. (1992). An intercomparison of methods for finding coupled patterns in climate data. *Journal of Climate*, 5(6), 541-560.
- von Storch, H. & Zwiers, F.W. (1999). *Statistical Analysis in Climate Research*. Cambridge University Press.
