---
title: "Untitled"
author: "Nick Gauthier"
date: "12/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidymodels)
library(scico)
library(stars)
```
```{r}
vfold_cv()
```

```{r}
set.seed(3527)
test_data <- data.frame(id = sort(sample(1:20, size = 80, replace = TRUE)))
test_data$dat <- runif(nrow(test_data))

set.seed(5144)
split_by_id <- group_vfold_cv(test_data, group = "id")

get_id_left_out <- function(x) unique(assessment(x)$id)

library(purrr)
table(map_int(split_by_id$splits, get_id_left_out))
#> 
#>  1  2  3  4  5  6  7  8  9 10 11 12 13 14 15 16 17 18 19 20 
#>  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1 

set.seed(5144)
split_by_some_id <- group_vfold_cv(test_data, group = "id", v = 7)
held_out <- map(split_by_some_id$splits, get_id_left_out)
table(unlist(held_out))
#> 
#>  1  2  3  4  5  6  7  8  9 10 11 12 13 14 15 16 17 18 19 20 
#>  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1  1 
# number held out per resample:
map_int(held_out, length)
```

```{r}
n <- 29
kfolds <- 5
  r <- n %% kfolds
  fold_times <- rep(n %/% kfolds, kfolds)
  fold_times[1:r] <- fold_times[1:r] + 1
 dat <- tibble(year = 1982:2010, fold = rep(1:kfolds, times = fold_times))
```

```{r}
dat_split <- group_vfold_cv(dat, group = fold, v = 5)

map(dat_split$splits, ~assessment(.x) %>% pull(year))
map(dat_split$splits, ~analysis(.x) %>% pull(year))

test_stars_rsample <- prism[,,,1:29] %>% st_set_dimensions('time', 1982:2010) 

# this works for subsetting the stars objects, although for those with disjoint dates you get a discrete time dimension
map(dat_split$splits, ~assessment(.x) %>% pull(year)) %>% map(~filter(test_stars_rsample, time %in% .x))
map(dat_split$splits, ~analysis(.x) %>% pull(year)) %>% map(~filter(test_stars_rsample, time %in% .x))
# this is actuall more or less what I'm doing already! so rsamples doesn't help much here
```

This time based resampling doesn't seem to work because its focused on the current step. better to do the vfold technique above
```{r}
dat_split_time <- dat %>%
  select(year) %>%
  sliding_window(assess_start = 1, assess_stop = 5, step = 6, complete = FALSE)

map(dat_split_time$splits, analysis)
map(dat_split_time$splits, assessment)
```



```{r}
prism <- readRDS('../inst/testdata/prism_test.RDS')

test <- prism %>%
  #split('time') %>%
  as_tibble() %>%
  group_by(x, y) %>%
  mutate(terms = as.character(cur_group_id()), .before = 1, .keep = 'unused') %>%
  ungroup()

indices <- select(test, terms, x, y) %>%
  unique()


test_prep <- test %>%
  select(-c(x, y)) %>%
  pivot_wider(names_from = terms, values_from = tmean) %>%
  units::drop_units() %>%
  recipe(~.) %>%
  update_role(time, new_role = 'id') %>%
  step_nzv(all_predictors()) %>%
 # step_normalize(all_predictors()) %>%
  step_center(all_predictors()) %>%
 step_pca(all_predictors(), id = 'pca') %>%
  step_scale(all_numeric_predictors()) %>%
  prep()

juice(test_prep) %>%
  pivot_longer(-time) %>%
  ggplot(aes(time, value)) +
  geom_line() +
  facet_wrap(~name)

tidy(test_prep, id = 'pca') %>% # this doesn't add 0's in the PC names so they come in the right order!
  filter(component %in% c('PC1', 'PC2', 'PC3', 'PC4', 'PC5')) %>%
  left_join(indices) %>%
  select(x, y, component, value, -c(terms, id)) %>%
  st_as_stars(dims = 1:3) %>%
  st_set_crs(st_crs(prism)) %>%
  plot



tidy(test_prep, id = 'pca', type = 'variance') %>% 
  filter(terms == 'percent variance', component <= 5) %>%
  ggplot(aes(component, value)) +
  geom_line()

```
so the step_pca process is nice for all the tidymodels interoperability reasons, but it does make it difficult for me to do things like monthly normalization, scaling the eofs and amplitudes how I want, etc.

either way, at least the above would be a good approach for doing pls or any other tidymodels method if I need to compare


```{r}
cera <- cera_dat %>%
    dplyr::mutate(SWE = SWE * sqrt(cos(y * pi / 180))) %>% # area weight, sqrt means covariance matrix weighted by cos(lat)
    tidyr::spread(year, SWE) %>%
    dplyr::select(-x, -y) %>%
    t() %>%
  as_tibble()

cera_space <- cera_dat %>%
    tidyr::spread(year, SWE) %>%
    dplyr::select(x, y) %>%
  mutate(id = paste0('V', 1:n()))

prism_pca <- recipe(prism) %>%
  update_role(everything(), new_role = 'predictor') %>%
  check_missing(everything()) %>%
  step_center(all_numeric()) %>%
  step_pca(all_numeric(), num_comp = 5, id = 'pca') %>%
  prep()

cera_pca <- recipe(cera) %>%
  update_role(everything(), new_role = 'outcome') %>%
  check_missing(everything()) %>%
  step_center(all_numeric()) %>%
  step_pca(all_numeric(), num_comp = 5, id = 'pca') %>%
  prep()


prism_pca
tidy(cera_pca, number = 3) 
```

This gets you the eigenvalues
```{r}
tidy(cera_pca, id = 'pca', type = 'variance') %>% filter(terms == 'percent variance') %>% pull(value) %>% .[1:10] %>% plot
```



```{r}
#get spatial ids

tidy(cera_pca, id = 'pca', type = 'coef') %>% filter(readr::parse_number(component) <= 5) %>% left_join(cera_space, by = c('terms' = 'id')) %>% ggplot(aes(x, y)) + geom_raster(aes(fill = value)) + coord_quickmap() + scale_fill_scico(palette = 'broc', name = 'Correlation') +
  facet_wrap(~component)

tidy(prism_pca, id = 'pca', type = 'coef') %>% filter(readr::parse_number(component) <= 5) %>% left_join(prism_space, by = c('terms' = 'id')) %>% ggplot(aes(x, y)) + geom_raster(aes(fill = value)) + coord_quickmap() + scale_fill_scico(palette = 'broc', name = 'Correlation') +
  facet_wrap(~component) + theme_void()
```

```{r}

```


```{r}
test_rec_ica <- recipe(dat %>% select(1:10000)) %>%
  step_center(all_numeric()) %>%
  step_ica(all_numeric(), num_comp = 5)

test_prep_ica <- prep(test_rec_ica, select(dat, 1:10000))
bake(test_prep_ica, new_data = NULL)

tidy(test_prep_ica, number = 2, type = 'coef') %>% filter(component %in% c('IC1', 'IC2', 'IC3', 'IC4', 'IC5')) %>% left_join(space, by = c('terms' = 'id')) %>% ggplot(aes(x, y)) + geom_raster(aes(fill = value)) + coord_quickmap() + scale_fill_scico(palette = 'broc', name = 'Correlation') +
  facet_wrap(~component)

```
```{r}
dat
```

