---
title: "vignette"
format: html
editor: visual
---

```{r}
# devtools::document()
# devtools::build(vignettes = FALSE)
# install.packages("~/Dropbox (UFL)/Projects/tidyEOF_0.1.0.tar.gz", repos = NULL, type = "source")
library(tidyEOF)
library(tidyverse)
library(stars)
prism <- system.file('testdata/prism_test.RDS', package = 'tidyEOF') %>%
  readRDS()
```

```{r}
clim <- get_climatology(prism)

clim
```

```{r}
ggplot() +
  geom_stars(data = clim) +
  scale_fill_viridis_c(option = 'magma') +
  facet_wrap(~statistic) +
  coord_quickmap() +
  theme_minimal()
```

```{r}
clim_mon <- get_climatology(prism, monthly = TRUE)

clim_mon
```

```{r}
ggplot() +
  geom_stars(data = clim_mon[,,,,1]) +
  scale_fill_viridis_c(option = 'magma') +
  facet_wrap(~month) +
  coord_quickmap() +
  theme_minimal()

ggplot() +
  geom_stars(data = clim_mon[,,,,2]) +
  scale_fill_viridis_c(option = 'magma') +
  facet_wrap(~month) +
  coord_quickmap() +
  theme_minimal()
```

```{r}
get_anomalies(prism, scale = TRUE, monthly = TRUE)
```

```{r}
plot_scree(prism)
plot_scree(prism, scale = TRUE)
plot_scree(prism, scale = FALSE, monthly = TRUE)
plot_scree(prism, scale = TRUE, monthly = TRUE)
```

```{r}
patterns <- get_patterns(prism, k = 2, monthly = TRUE, scale = FALSE, rotate = FALSE)

plot(patterns)
plot_eofs(patterns, scaled = TRUE, rawdata = get_anomalies(prism, monthly = TRUE)) # seems wrong, ah, its because of the monthly mean standardization! so this should instead do something with the scaling . . .
```

```{r}
patterns
summary(patterns)

plot(patterns)
plot(patterns, 'amplitudes')
plot(patterns, 'scree')


get_patterns(prism, k = 2, monthly = FALSE, scale = TRUE) %>%
  plot()
get_patterns(prism, k = 2, monthly = FALSE, scale = FALSE) %>%
  plot()


get_patterns(prism, k = 2, monthly = TRUE, scale = TRUE) %>%
  plot()
get_patterns(prism, k = 2, monthly = TRUE, scale = FALSE) %>%
  plot()

get_patterns(prism, k = 2, monthly = FALSE, scale = TRUE, rotate = TRUE) %>%
  plot()
get_patterns(prism, k = 2, monthly = FALSE, scale = FALSE, rotate = TRUE) %>%
  plot()


get_patterns(prism, k = 2, monthly = TRUE, scale = TRUE, rotate = TRUE) %>%
  plot()
get_patterns(prism, k = 2, monthly = TRUE, scale = FALSE, rotate = TRUE) %>%
  plot()
```

```{r}
plot_amps(patterns)
plot_amps(patterns, scaled = FALSE)
```

```{r}
get_patterns(prism, k = 2, monthly = TRUE, scale = TRUE) %>%
  plot_amps(scaled = TRUE)
get_patterns(prism, k = 2, monthly = TRUE, scale = TRUE) %>%
  plot_amps(scaled = FALSE)
```

```{r}
get_patterns(prism, k = 2, monthly = FALSE, scale = TRUE) %>%
  plot_amps(scaled = TRUE)
get_patterns(prism, k = 2, monthly = FALSE, scale = TRUE) %>%
  plot_amps(scaled = FALSE)
get_patterns(prism, k = 2, monthly = FALSE, scale = FALSE) %>%
  plot_amps(scaled = TRUE)
get_patterns(prism, k = 2, monthly = FALSE, scale = FALSE) %>%
  plot_amps(scaled = FALSE)
```

and here having monthly TRUE does something to the dimensions! this could be a test

```{r}
t1 <- get_patterns(filter(prism), scale = FALSE, monthly = FALSE, rotate = FALSE, k =2) %>% reconstruct_field(nonneg = FALSE) 

t2 <- get_patterns(filter(prism), scale = FALSE, monthly = TRUE, rotate = FALSE, k =2) %>% reconstruct_field(nonneg = FALSE) 

t1 - filter(prism)
t2 - filter(prism)
stars::st_dimensions(t2)
stars::st_dimensions(t1)

t1 <- get_anomalies(prism, monthly = FALSE)
t2 <- get_anomalies(prism, monthly = TRUE)
library(stars)
d1 = st_dimensions(t1)
	d2 = st_dimensions(t2)
	crs1 = st_crs(d1)
	crs2 = st_crs(d2)
	st_crs(d1) = st_crs(NA)
	st_crs(d2) = st_crs(NA)
	n = min(length(d1), length(d2))
	isTRUE(all.equal(d1[1:n], d2[1:n], check.attributes = FALSE)) && crs1 == crs2
d1$time$values
d2$time$values
## get anomalies reshuffles time! it gives it by month rather than by year!!!
t_order <- d2$time$values %>% order
t2[,,,t_order] %>% plot
time(t2) %>% order
t2 %>% slice(., 'time', order(time(.))) %>% plot
t2[,,,]

recon_test <- map(1:10, ~get_patterns(filter(prism), scale = TRUE, monthly = TRUE, rotate = TRUE, k =.x) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism))  %>% stars::st_apply(1:2, function(x) sqrt(mean(x^2)))) %>%
  append(list(along = 'k')) %>%
  do.call(c, .)
  
  get_patterns(filter(prism), scale = TRUE, monthly = TRUE, k = 4)$climatology %>% filter(month == month.name[1])
  
    get_patterns(filter(prism), scale = TRUE, monthly = TRUE, k = 4)$amplitudes$time %>% class
```

```{r}
get_climatology(filter(prism), monthly = TRUE)

time_series_redim <- stars::st_dimensions(x = stars::st_dimensions(prism)$x,
                                           y = stars::st_dimensions(prism)$y,
                                           year = unique(lubridate::year(time(prism))),
                                           month = month.name) %>%
  st_redimension(prism, .)
```

so this error suggests that something that sweep_months is doing is breaking the output in such a way that later calls to filter are bad

```{r}
test <- get_patterns(filter(prism), scale = TRUE, monthly = TRUE, k = 4) %>% 
  reconstruct_field(nonneg = FALSE)

test_clim  <- get_patterns(filter(prism), scale = TRUE, monthly = TRUE, k = 4)$climatology %>% slice('var', 1) %>% units::drop_units()


filter(test, lubridate::month(time) == 1)
filter(prism, lubridate::month(time) == 1)
st_dimensions(test)$time$values %>% class
st_dimensions(prism)$time$values %>% class

test - filter(prism)
test$tmean %>% dim

time(test)
slice(prism, 'time', 1:2) %>% st_dimensions() %>% .$time %>% .$values
get_anomalies(prism, scale = TRUE, monthly = TRUE)  %>% st_dimensions() %>% .$time %>% .$values
# so sweep stars gives you an interval object every time!

filter(prism, lubridate::month(time) == 1) %>% time

get_patterns(filter(prism), scale = FALSE, monthly = TRUE, k = 4) %>% 
  reconstruct_field(nonneg = FALSE) %>% st_dimensions() %>% .$time %>% .$values
 
sweep_months(test, test_clim, '+')

 purrr::map(1:12, ~ FUN(filter(e1, lubridate::month(time) == .x), abind::adrop(filter(e2, month == month.name[.x])))) %>%
    do.call(c, .) %>%
    slice(., 'time', order(time(.)))  
 
 filter(test, lubridate::month(time) == 12)
 time(test) %>% lubridate::month()
 abind::adrop(filter(test_clim, month == month.name[1]))
 
 
 purrr::map(1:12, ~ `+`(filter(e1, lubridate::month(time) == .x), abind::adrop(filter(e2, month == month.name[.x])))) %>%
    do.call(c, .) %>%
    slice(., 'time', order(time(.))) 
```

```{r}
recon_test1 <- map(1:20, ~get_patterns(filter(prism), scale = TRUE, monthly = TRUE, rotate = TRUE, k =.x) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism))  %>% stars::st_apply(1:2, function(x) sqrt(mean(x^2)))) %>%
  append(list(along = 'k')) %>%
  do.call(c, .)
recon_test2 <- map(1:20, ~get_patterns(filter(prism), scale = TRUE, monthly = TRUE, rotate = FALSE, k =.x) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism))  %>% stars::st_apply(1:2, function(x) sqrt(mean(x^2)))) %>%
  append(list(along = 'k')) %>%
  do.call(c, .)
recon_test3 <- map(1:20, ~get_patterns(filter(prism), scale = TRUE, monthly = FALSE, rotate = FALSE, k =.x) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism))  %>% stars::st_apply(1:2, function(x) sqrt(mean(x^2)))) %>%
  append(list(along = 'k')) %>%
  do.call(c, .)
recon_test4 <- map(1:20, ~get_patterns(filter(prism), scale = FALSE, monthly = FALSE, rotate = FALSE, k =.x) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism))  %>% stars::st_apply(1:2, function(x) sqrt(mean(x^2)))) %>%
  append(list(along = 'k')) %>%
  do.call(c, .)
recon_test5 <- map(1:20, ~get_patterns(filter(prism), scale = FALSE, monthly = TRUE, rotate = FALSE, k =.x) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism))  %>% stars::st_apply(1:2, function(x) sqrt(mean(x^2)))) %>%
  append(list(along = 'k')) %>%
  do.call(c, .)
recon_test6 <- map(1:20, ~get_patterns(filter(prism), scale = FALSE, monthly = TRUE, rotate = TRUE, k =.x) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism))  %>% stars::st_apply(1:2, function(x) sqrt(mean(x^2)))) %>%
  append(list(along = 'k')) %>%
  do.call(c, .)
recon_test7 <- map(1:20, ~get_patterns(filter(prism), scale = TRUE, monthly = FALSE, rotate = TRUE, k =.x) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism))  %>% stars::st_apply(1:2, function(x) sqrt(mean(x^2)))) %>%
  append(list(along = 'k')) %>%
  do.call(c, .)
```

```{r}
ggplot() +
  geom_stars(data = recon_test) +
  facet_wrap(~k) +
  scale_fill_viridis_c() +
  coord_quickmap()
```

```{r}
out <- bind_rows(
stars::st_apply(recon_test1, 3, mean) %>% as_tibble(),
stars::st_apply(recon_test2, 3, mean) %>% as_tibble(),
stars::st_apply(recon_test3, 3, mean) %>% as_tibble(),
stars::st_apply(recon_test4, 3, mean) %>% as_tibble(),
stars::st_apply(recon_test5, 3, mean) %>% as_tibble(),
stars::st_apply(recon_test6, 3, mean) %>% as_tibble(),
stars::st_apply(recon_test7, 3, mean) %>% as_tibble(), .id = 'rep')

out %>%
    group_by(rep) %>%
  summarise(test = min(mean)) %>%
  arrange(test)

ggplot(out, aes(k, mean, color = rep)) +
  geom_line()

```

so rotation and stuff is identical, good!

5-6 the best with monthly centering but no scaling

7-3 second with scale true but no monthly

4 basically the same with no scale no monthly

2 and 1 worse with scale true and monthly true -- why is this -- scaling removes variability?

The below errors should be 0 right?? is it because of of scaling amplitudes -- must be right?

still off by machine precision?

```{r}
get_patterns(filter(prism), monthly = TRUE,  k =36) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism)) %>% as_tibble() %>% summarise(sqrt(mean(tmean^2)))
get_patterns(filter(prism), monthly = FALSE,  k =36) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism)) %>% as_tibble() %>% summarise(sqrt(mean(tmean^2)))
get_patterns(filter(prism), monthly = TRUE, scale = TRUE, k =36) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism)) %>% as_tibble() %>% summarise(sqrt(mean(tmean^2)))
get_patterns(filter(prism), monthly = FALSE, scale = TRUE, k =36) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism)) %>% as_tibble() %>% summarise(sqrt(mean(tmean^2)))
```

now do that again with and without the area weight to see if that makes a difference for scale. it seems like there may be a minor impact of this!

```{r}
get_patterns(filter(prism), k =36) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism)) %>% as_tibble() %>% summarise(sqrt(mean(tmean^2)))
get_patterns(filter(prism), scale = TRUE, k =36) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism)) %>% as_tibble() %>% summarise(sqrt(mean(tmean^2)))

get_patterns(filter(prism), k =36,weight = FALSE) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism)) %>% as_tibble() %>% summarise(sqrt(mean(tmean^2)))
get_patterns(filter(prism), scale = TRUE, k =36, weight = FALSE) %>% reconstruct_field(nonneg = FALSE) %>% `-`(filter(prism)) %>% as_tibble() %>% summarise(sqrt(mean(tmean^2)))
```

```{r}
get_patterns(filter(prism), k =4) %>% plot_amps() 
```

```{r}
get_patterns(filter(prism), k =4, monthly = TRUE) %>% plot_eofs()
get_patterns(filter(prism), k =4, monthly = FALSE) %>% plot_eofs(scaled = TRUE, prism)
get_patterns(filter(prism), k =4, monthly = TRUE, scale = TRUE) %>% plot_eofs(scaled = TRUE, prism)

get_patterns(filter(prism), k =4, monthly = TRUE, rotate = TRUE) %>% plot_eofs(scaled = TRUE, prism)
```

```{r}
patterns$eofs %>% plot
```

## Coupled patterns

```{r}
coarse <- st_bbox(prism) %>%
  st_as_stars(dx = 0.25) %>%
  st_warp(prism, ., use_gdal = TRUE, method = 'average') %>%
  st_set_dimensions('band', names = 'time', values= st_get_dimension_values(prism, 'time')) %>%
  setNames('tmean') %>%
  mutate(tmean = units::set_units(tmean, degree_C))
```

```{r}
plot(coarse)
```

```{r}
get_patterns(coarse, k = 2, monthly = TRUE) %>% plot_eofs
```

```{r}
delta_add(prism, prism, monthly = TRUE) %>% plot
```

Comparing delta to CCA methods

```{r}
(delta_add(coarse, prism, monthly = TRUE) - filter(prism)) %>%
  units::drop_units() %>%
  pull(tmean) %>%
  `^`(2) %>%
  mean() %>%
  sqrt()
(delta_add(coarse, prism, monthly = FALSE) - (prism)) %>%
  units::drop_units() %>%
  pull(tmean) %>%
  `^`(2) %>%
  mean() %>%
  sqrt()

(predict_cca(get_patterns(coarse, k = 5), get_patterns(filter(prism), k = 5), coarse, k = 5) - filter(prism)) %>%
  units::drop_units() %>%
  pull(tmean) %>%
  `^`(2) %>%
  mean() %>%
  sqrt()

(predict_cca(get_patterns(coarse, k = 5, monthly = TRUE), get_patterns(filter(prism), k = 5, monthly = TRUE), coarse, k = 5) - filter(prism)) %>%
  units::drop_units() %>%
  pull(tmean) %>%
  `^`(2) %>%
  mean() %>%
  sqrt()

(predict_cca(get_patterns(coarse, k = 5, monthly = TRUE, scale = TRUE), get_patterns(filter(prism), k = 5, monthly = TRUE, scale = TRUE), coarse, k = 5) - filter(prism)) %>%
  units::drop_units() %>%
  pull(tmean) %>%
  `^`(2) %>%
  mean() %>%
  sqrt()

(predict_cca(get_patterns(coarse, k = 5, monthly = FALSE, scale = TRUE), get_patterns(filter(prism), k = 5, monthly = FALSE, scale = TRUE), coarse, k = 5) - filter(prism)) %>%
  units::drop_units() %>%
  pull(tmean) %>%
  `^`(2) %>%
  mean() %>%
  sqrt()
```

```{r}
map_dbl(1:36, ~(predict_cca(get_patterns(coarse, k = .x, monthly = TRUE), get_patterns(filter(prism), k = .x, monthly = TRUE), coarse, k = .x) - filter(prism)) %>%
  units::drop_units() %>%
  pull(tmean) %>%
  `^`(2) %>%
  mean() %>%
  sqrt())
```

```{r}
ggplot() +
  geom_stars(data = (predict_cca(get_patterns(coarse, k = 5, monthly = TRUE), get_patterns(filter(prism), k = 5, monthly = TRUE), coarse, k = 5) )) +
  facet_wrap(~time) +
  coord_quickmap() +
  scale_fill_viridis_c(option = 'magma')

ggplot() +
  geom_stars(data = prism) +
  facet_wrap(~time) +
  coord_quickmap() +
  scale_fill_viridis_c(option = 'magma')

ggplot() +
  geom_stars(data = (predict_cca(get_patterns(coarse, k = 5, monthly = TRUE), get_patterns(filter(prism), k = 5, monthly = TRUE), coarse, k = 5) ) - filter(prism)) +
  facet_wrap(~time) +
  coord_quickmap() +
  scale_fill_distiller(palette = 'RdBu')


ggplot() +
  geom_stars(data = (delta_add(coarse, prism, monthly = TRUE) - filter(prism)) ) +
  facet_wrap(~time) +
  coord_quickmap() +
  scale_fill_distiller(palette = 'RdBu')
```

This shows the machine precision issues with values not exactly equal to zero after centering - - - this is ok in the real world, but does it make any difference for the svd?

```{r}
prism %>% get_anomalies(monthly = TRUE) %>% st_apply(1:2, mean) %>% plot
prism %>% get_anomalies(monthly = FALSE) %>% st_apply(1:2, mean) %>% plot
(units::drop_units(prism) - st_apply(prism, 1:2, mean)) %>% st_apply(1:2, mean) %>% plot
```

```{r}
delta_mul(prism, prism, monthly = FALSE) %>% plot
```
