---
title: "Introduction to tidyeof"
format: html
editor: visual
---

## Overview

The `tidyeof` package provides tools for analyzing spatiotemporal climate data using Empirical Orthogonal Functions (EOFs), also known as Principal Component Analysis (PCA). This vignette will:

1.  Introduce the concept of EOF analysis
2.  Demonstrate how to use the package
3.  Show how to interpret and validate results
4.  Provide examples of common analysis workflows

## What are EOFs?

EOFs are a powerful technique for analyzing spatiotemporal climate data. They help us identify dominant spatial patterns of variability and how these patterns evolve over time. Mathematically, EOFs decompose a spatiotemporal field into:

1.  Spatial patterns (EOFs or loading patterns)
2.  Time series (Principal Components or pattern amplitudes)
3.  Explained variance (eigenvalues)

Consider a temperature anomaly field over North America. An EOF analysis might reveal that: - The first EOF shows a continent-wide warming/cooling pattern - The second EOF might show a dipole between east and west - Each pattern has an associated time series showing when it was active - The eigenvalues tell us how much of the total variance each pattern explains

## Getting Started

First, let's load the package and some example data:

```{r}
library(tidyeof)
library(stars)
library(tidyverse)
library(ggplot2)

# Load example data (monthly temperature anomalies)
temp_anomalies <- system.file('testdata/prism_test.RDS', package = 'tidyeof') |>
  readRDS()
```

## Basic EOF Analysis

The main function is `patterns()`:

```{r}
pat <- patterns(temp_anomalies, k = 3) # k selects number of patterns
```

The resulting `patterns` object contains: - Spatial patterns (`pat$eofs`) - Time series (`pat$amplitudes`) - Explained variance (`pat$eigenvalues`) - Processing metadata

```{r}
get_climatology(temp_anomalies) |> plot()

get_climatology(temp_anomalies, monthly = TRUE) |> plot()
```

## Visualizing Results

The package provides several plotting methods:

```{r}
plot(pat)
```

```{r}
screeplot(pat)
```

```{r}
patterns(temp_anomalies, k = 3, rotate = TRUE, scale = TRUE) |> plot()
```

```{r}
patterns(temp_anomalies, k = 3, rotate = TRUE) |> plot()
```

```{r}
p1 <- plot(pat, 'eofs') + facet_wrap(~PC, nrow = 1) + theme_minimal() + labs(x='', y='')
p2 <- plot(pat, 'amplitudes') + facet_wrap(~PC, nrow = 1)
(p1) / (p2) + plot_layout(heights = c(5, 1), guides = 'collect')

(p1) + (p2) + plot_layout(heights = c(5, 1), guides = 'collect')
```

```{r}
p3 <- plot(pat, 'eofs') + facet_wrap(~PC, ncol = 1) + labs(x='', y='')
p4 <- plot(pat, 'amplitudes') + facet_wrap(~PC, ncol = 1)
(p3) | (p4) + plot_layout(widths = c(10, 1), guides = 'collect')

```

```{r}
# Plot spatial patterns and amplitudes
plot(pat)
```

## Understanding the Output

Let's break down what each component means:

### Spatial Patterns (EOFs)

```{r}
pat
# Look at first EOF
pat$eofs  # Using $ method
pat[1]    # Using [ method
plot(pat$eofs)

pat$eofs
plot(pat[3])
```

The EOF patterns show spatial loadings where: - Red areas move together in one direction - Blue areas move together in the opposite direction - Darker colors indicate stronger relationships - White areas have little relationship with the pattern

### Time Series (PCs)

```{r}
# Extract PC time series
pat$amplitudes

# Plot amplitudes
plot(pat, type = "amplitudes")
```

The time series show:

-   When each pattern was active (positive values)
-   When it was in its opposite phase (negative values)
-   The magnitude of the pattern at each time

### Explained Variance

```{r}
summary(pat)
```

This shows:

-   How much variance each pattern explains

-   Cumulative variance explained

-   Statistical significance of patterns

## Advanced Topics

### Rotation

EOF patterns are mathematically constrained to be orthogonal, which may not reflect physical reality. Rotation can help find more physically interpretable patterns:

```{r}
rotated_pat <- patterns(temp_anomalies,
                        k = 4,
                        rotate = TRUE)

plot(rotated_pat)
```

### Pattern Significance

Not all patterns are statistically significant. We can test significance using: - North's Rule of Thumb - Monte Carlo methods - The scree plot

### Latitude Weighting

For global analyses, grid cells have different areas at different latitudes. The `weight` parameter applies area weighting to account for this.

## Common Workflows

### Finding Dominant Patterns

```{r}
# Get patterns, keeping enough to explain 90% of variance
var_explained <- cumsum(pat$eigenvalues$variance) /
                 sum(pat$eigenvalues$variance)
k_needed <- which(var_explained > 0.9)[1]

dominant_pat <- patterns(temp_anomalies, k = k_needed)
```

### Seasonal Analysis

```{r}
# Get patterns with monthly climatology
seasonal_pat <- patterns(temp_anomalies, monthly = TRUE)

# Look at winter patterns
winter_months <- pat$amplitudes$time |>
  lubridate::month() %in% c(12, 1, 2)
winter_amplitudes <- pat$amplitudes[winter_months, ]
```

### Pattern Reconstruction

```{r}
# Reconstruct field using patterns
reconstructed <- reconstruct(pat)

# Calculate reconstruction error
error <- temp_anomalies - reconstructed
```

## Best Practices

1.  **Data Preparation**
    -   Remove seasonal cycle if not studying seasonal patterns
    -   Consider whether to standardize the field
    -   Check for and handle missing values
2.  **Pattern Selection**
    -   Use scree plot to guide pattern retention
    -   Consider physical interpretability
    -   Test pattern significance
3.  **Interpretation**
    -   Compare patterns to known physical modes
    -   Consider the timescales of variability
    -   Remember patterns are statistical constructs

## Common Issues

1.  **Scale Mixing**
    -   EOFs can mix processes operating at different scales
    -   Consider scale separation techniques if needed
2.  **Domain Dependence**
    -   Patterns depend on domain choice
    -   Choose domain based on physical understanding
3.  **Orthogonality Constraint**
    -   Patterns are mathematically, not physically, constrained
    -   Consider rotation for physical interpretation

## References

For more information on EOF/PCA analysis in climate science:

1.  Wilks, D.S. (2011). Statistical Methods in the Atmospheric Sciences
2.  von Storch, H. & Zwiers, F.W. (1999). Statistical Analysis in Climate Research
3.  Hannachi, A., et al. (2007). Pattern hunting in climate: a new method...

## Next Steps

-   Try the examples with your own data
-   Explore advanced options like rotation
-   Check the function documentation for details
-   See the package website for more examples
